---
title: "SpotSeq Dataset Analysis"
author: "SpotSeq Analysis Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(readxl)
library(janitor)
library(openxlsx)
library(gt)
library(glue)

data_path      <- Sys.getenv("DATA_PATH")
drive_folder_id <- Sys.getenv("DRIVE_FOLDER_ID")

stopifnot(
  "DATA_PATH not set in .Renviron"      = nchar(data_path) > 0,
  "DRIVE_FOLDER_ID not set in .Renviron" = nchar(drive_folder_id) > 0
)
```

```{r load-data}
#| label: load-data

# Read all columns as text first to capture displayed cell values, not formulas
raw <- read_excel(
  file.path(data_path, "SpotSeq_Project_Dataset_deidentified_2.1.xlsx"),
  col_types = "text"
)

df_raw    <- raw |> clean_names()
df        <- df_raw |> filter(!is.na(patient_id) & patient_id != "")
n_raw     <- nrow(raw)
n_spacer  <- n_raw - nrow(df)
n_records <- nrow(df)
n_pts     <- n_distinct(df$patient_id)
```

```{r specimen-standardise}
#| label: specimen-standardise

blood_vals <- toupper(c("Blood","Blood, venous","Lesional blood"))
lymph_vals <- toupper(c(
  "Lymphatic fluid","Lymphatic fluid, Site A","Lymphatic fluid, Site B",
  "Lymphatic fluid from left groin","Lymphatic cyst fluid",
  "Lymphatic cyst fluid from lymph node","Cyst fluid",
  "Lymphatic Malformation","Lymphatic malformation",
  "Chest, Lymphatic malformation"
))
other_vals <- toupper(c(
  "Aspirated Cyst Fluid","aspirated cyst fluid",
  "Abdomen, Lymphatic Cyst Fluid","Ankle, Left","Neck",
  "Tracheocutaneous fistula","Peritoneal fluid"
))

df <- df |>
  mutate(
    specimen_source_raw = toupper(trimws(specimen_source)),
    specimen_source_std = case_when(
      specimen_source_raw %in% blood_vals ~ "Blood",
      specimen_source_raw %in% lymph_vals ~ "Lymphatic Fluid",
      specimen_source_raw %in% other_vals ~ "Other",
      TRUE                                ~ NA_character_
    )
  )
```

```{r classify-results}
#| label: classify-results

df <- df |>
  mutate(
    # Standardise result: trim, upper
    result_clean   = toupper(trimws(result_pos_neg)),
    status_clean   = toupper(trimws(status)),

    # Classification flags
    is_qns        = status_clean %in% c("QNS","QNS FOR REPEAT","REPEAT") |
                    result_clean %in% c("QNS","QNS FOR REPEAT"),
    is_cancelled  = status_clean %in% c("CANCELLED","CANCELED") |
                    result_clean %in% c("CANCELLED","CANCELED"),

    # Blank result entries — omit from analysis
    is_blank      = is.na(result_pos_neg) | trimws(result_pos_neg) == "",

    # Valid = has a POS or NEG result, not QNS/cancelled/blank
    # status_clean == "COMPLETE" NOT required — some valid results have NA status
    is_valid      = !is_qns & !is_cancelled & !is_blank &
                    !is.na(result_clean) & result_clean != "" &
                    (grepl("^POS", result_clean) | grepl("^NEG", result_clean)),

    # Positive / Negative: NEG* and any capitalisation variant treated as NEG
    is_pos        = is_valid & grepl("^POS", result_clean),
    is_neg        = is_valid & grepl("^NEG", result_clean),   # catches NEG, NEG*, Neg, neg, etc.

    assay_std     = toupper(trimws(assay_ordered)),

    # Age: use displayed value from column 42 (age_at_date_of_collection_in_days)
    # Divide by 365.25 for decimal years — handles infants correctly
    age_days      = suppressWarnings(as.numeric(age_at_date_of_collection_in_days)),
    age_years     = age_days / 365.25,

    # Sex
    sex_val       = assigned_sex_at_birth,

    # Specimen type standardised (col 6)
    specimen_type_std = case_when(
      toupper(trimws(specimen_type)) == "BLOOD"     ~ "Blood",
      grepl("CYST", toupper(trimws(specimen_type))) ~ "Cyst Fluid",
      TRUE                                          ~ "Other"
    ),

    # Internal vs external patient (col 43)
    internal_val  = toupper(trimws(internal_patient_y_n)),

    # Volume / DNA / assay metrics — read from displayed values
    vol_val       = suppressWarnings(as.numeric(fluid_volume_extracted_converted)),
    dna_conc_val  = suppressWarnings(as.numeric(
                      dna_stock_ng_ul_blood_0_185ng_ul_reject_cyst_fluid_0_37ng_ul_reject)),
    dna_used_val  = suppressWarnings(as.numeric(total_ng_dna_used_in_assay)),
    ex21_val      = suppressWarnings(as.numeric(exon_21_wt_1500_avg_number_of_droplets)),
    ex10_val      = suppressWarnings(as.numeric(exon_10_wt_1500_avg_number_droplets)),

    # Table 4 — Clinical impact columns (44–49), exact clean_names() output
    mgmt_change        = toupper(trimws(any_documented_management_change_in_management_post_testing_y_n)),
    therapy_initiated  = toupper(trimws(targeted_therapy_initiated_post_testing_e_g_alpelisib_sirolimus_y_n)),
    therapy_modified   = toupper(trimws(existing_therapy_modified_or_dose_adjusted_post_testing_y_n_na)),
    procedure_changed  = toupper(trimws(surgical_or_procedural_plan_changed_post_testing_y_n_na)),
    additional_testing = toupper(trimws(patient_underwent_additional_testing_based_on_spot_seq_results_y_n_van)),
    mgmt_unavailable   = toupper(trimws(management_documentation_post_testing_unavailable_not_reported_y_n_na))
  )

# Assign each record to exactly one mutually exclusive category
# Priority: valid > qns > cancelled > unclassified
df <- df |>
  mutate(
    record_category = case_when(
      is_valid     ~ "valid",
      is_qns       ~ "qns",
      is_cancelled ~ "cancelled",
      is_blank     ~ "blank",
      TRUE         ~ "unclassified"
    )
  )

# Dynamically resolve column names that may vary between dataset versions
find_col <- function(df, pattern, label) {
  hits <- grep(pattern, names(df), value = TRUE, ignore.case = TRUE)
  if (length(hits) == 0) {
    message("WARNING: No column matching '", pattern, "' for ", label, " — will be NA.")
    return(NULL)
  }
  if (length(hits) > 1)
    message("NOTE: Multiple matches for '", pattern, "': ", paste(hits, collapse=", "),
            " — using: ", hits[1])
  hits[1]
}
col_single_probe <- find_col(df_raw, "single.probe",  "single_probe_val")
col_gene         <- find_col(df_raw, "^gene$",         "gene_val")
col_variant      <- find_col(df_raw, "^variant$",      "variant_val")

df <- df |>
  mutate(
    single_probe_val = if (!is.null(col_single_probe))
                         suppressWarnings(as.numeric(.data[[col_single_probe]]))
                       else NA_real_,
    gene_val    = if (!is.null(col_gene))    toupper(trimws(.data[[col_gene]]))    else NA_character_,
    variant_val = if (!is.null(col_variant)) toupper(trimws(.data[[col_variant]])) else NA_character_
  )

# Filtered subsets — strictly mutually exclusive
df_valid    <- df |> filter(record_category == "valid")
df_qns      <- df |> filter(record_category == "qns")
df_cancel   <- df |> filter(record_category == "cancelled")
# Omit blanks from all counts
df_analysed <- df |> filter(record_category != "blank")
```

```{r data-reconciliation}
#| label: data-reconciliation

# Explicit accounting of all records — surfaces any unclassified entries
n_total    <- nrow(df_analysed)
n_valid    <- nrow(df_valid)
n_qns_n    <- nrow(df_qns)
n_can      <- nrow(df_cancel)
n_blanks   <- sum(df$record_category == "blank", na.rm = TRUE)
n_other    <- sum(df$record_category == "unclassified", na.rm = TRUE)
# All counts are strictly mutually exclusive via record_category
# n_valid + n_qns_n + n_can + n_other should equal n_total

tibble(
  Category = c(
    "Total rows in raw file",
    "Spacer rows excluded (no Patient ID)",
    "Records retained (with Patient ID)",
    "  Blank result entries (omitted from analysis)",
    "  Records included in analysis",
    "    Valid (analyzable)",
    "    QNS",
    "    Cancelled",
    "    Unclassified (investigate if > 0)"
  ),
  Count = c(
    n_raw, n_spacer, n_records,
    n_blanks, n_total,
    n_valid, n_qns_n, n_can, n_other
  )
) |>
  gt() |>
  tab_header(title = "Data Reconciliation — Record Accounting") |>
  tab_options(table.width = pct(50), column_labels.font.weight = "bold")
```

```{r patient-level}
#| label: patient-level

# Deduplicate to one row per unique patient using sample_id_for_di_dataset
# Use first test entry per patient — sort by row order (as read from Excel)
# since collection_date is stored as text and row order reflects entry order

df_patients <- df_valid |>
  mutate(row_order = row_number()) |>
  group_by(patient_id) |>
  slice_min(row_order, n = 1, with_ties = FALSE) |>
  ungroup()

n_pts_valid <- n_distinct(df_patients$patient_id)


```

```{r helpers}
#| label: helpers

fmt_ms  <- function(m, s) sprintf("%.1f (%.1f)", m, s)
fmt_med <- function(med, lo, hi) sprintf("%.1f [%.1f–%.1f]", med, lo, hi)
fmt_np  <- function(n, p) sprintf("%d (%.1f%%)", as.integer(n), p)
cell2   <- function(n, p) sprintf("%d (%.1f%%)", as.integer(n), p)
iqr_s   <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return("—")
  sprintf("%.1f [%.1f–%.1f]", median(x), quantile(x,.25), quantile(x,.75))
}
msd_s   <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return("—")
  sprintf("%.1f (%.1f)", mean(x), sd(x))
}
fmt_p <- function(p) {
  if (is.na(p) || !is.finite(p)) return("--")
  if (p < 0.001) return("<0.001")
  sprintf("%.3f", p)
}
safe_wilcox <- function(x, y) {
  x <- x[!is.na(x)]; y <- y[!is.na(y)]
  if (length(x) < 2 || length(y) < 2) return(NA_real_)
  tryCatch(wilcox.test(x, y, exact = FALSE)$p.value, error = function(e) NA_real_)
}
safe_fisher <- function(grp, flag) {
  tab <- table(grp, flag)
  if (any(dim(tab) < 2)) return(NA_real_)
  tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
}
spec_bk <- function(s, nn) {
  enc <- df_valid |>
    filter(specimen_source_std == s) |>
    group_by(patient_id, collection_date) |>
    summarise(types = n_distinct(specimen_source_std), .groups = "drop") |>
    filter(types > 1) |> nrow()
  bl   <- sum(df_valid$specimen_source_std == s & df_valid$is_pos, na.rm = TRUE)
  ly   <- sum(df_valid$specimen_source_std == s & df_valid$is_pos, na.rm = TRUE)
  both <- enc
  n_bl <- sum(df_valid$spec_grp == "Blood only",           na.rm = TRUE)
  n_ly <- sum(df_valid$spec_grp == "Lymphatic fluid only", na.rm = TRUE)
  list(
    blood  = fmt_np(n_bl, n_bl / nn * 100),
    lymph  = fmt_np(n_ly, n_ly / nn * 100),
    both   = fmt_np(both, both / nn * 100)
  )
}
```

```{r table1-calcs}
#| label: table1-calcs

# Patient-level stats use df_patients (first test per unique sample_id_for_di_dataset)

# Age summary — using decimal years from AP column
age_n      <- sum(!is.na(df_patients$age_years))
age_mean   <- mean(df_patients$age_years,   na.rm = TRUE)
age_sd     <- sd(df_patients$age_years,     na.rm = TRUE)
age_med    <- median(df_patients$age_years, na.rm = TRUE)
age_q1     <- quantile(df_patients$age_years, 0.25, na.rm = TRUE)
age_q3     <- quantile(df_patients$age_years, 0.75, na.rm = TRUE)
age_min    <- min(df_patients$age_years,    na.rm = TRUE)
age_max    <- max(df_patients$age_years,    na.rm = TRUE)

# Pediatric / Adult split (< 18 years)
n_peds     <- sum(df_patients$age_years < 18,  na.rm = TRUE)
n_adult    <- sum(df_patients$age_years >= 18, na.rm = TRUE)
pct_peds   <- n_peds  / age_n * 100
pct_adult  <- n_adult / age_n * 100

# Sex
sex_clean  <- toupper(trimws(df_patients$sex_val))
n_female   <- sum(sex_clean == "F", na.rm = TRUE)
sex_n      <- sum(!is.na(df_patients$sex_val))
n_sex_n    <- sum(is.na(df_patients$sex_val))
pct_female <- n_female / sex_n * 100

# Internal vs External
internal_clean <- toupper(trimws(df_patients$internal_val))
n_internal     <- sum(internal_clean == "Y", na.rm = TRUE)
n_external     <- sum(internal_clean == "N", na.rm = TRUE)
n_int_ext      <- sum(!is.na(df_patients$internal_val) & df_patients$internal_val != "")
pct_int        <- if(n_int_ext > 0) n_internal / n_int_ext * 100 else 0
pct_ext        <- if(n_int_ext > 0) n_external / n_int_ext * 100 else 0
```

---

# Table 1 — Patient and Cohort Characteristics

```{r table1-display}
#| label: table1-display

t1 <- tibble(
  Characteristic = c(
    "Demographics",
    "  Age at collection, years",
    "    Mean (SD)",
    "    Median [IQR]",
    "    Range",
    "  Age group",
    "    Pediatric (< 18 years)",
    "    Adult (≥ 18 years)",
    "  Female sex assigned at birth, n (%)",
    "Patient Origin",
    "  Internally processed, n (%)",
    "  Externally referred, n (%)",
    "Testing Overview",
    "  Total tests ordered",
    "  Unique patients",
    "  Internally processed tests, n (%)"
  ),
  `N (%) / Value` = c(
    "",
    "",
    fmt_ms(age_mean, age_sd),
    fmt_med(age_med, age_q1, age_q3),
    sprintf("%.2f – %.1f", age_min, age_max),
    "",
    fmt_np(n_peds,  pct_peds),
    fmt_np(n_adult, pct_adult),
    glue("{fmt_np(n_female, pct_female)} (N = {sex_n} of {n_pts_valid} patients)"),
    "",
    fmt_np(n_internal, pct_int),
    fmt_np(n_external, pct_ext),
    "",
    as.character(n_records),
    as.character(n_pts),
    fmt_np(n_internal, pct_int)
  )
)

section_rows <- which(t1$Characteristic %in%
                        c("Demographics","Patient Origin","Testing Overview"))

gt_t1 <- t1 |>
  gt() |>
  tab_header(
    title = glue("Table 1. Patient and Cohort Characteristics (N = {n_records} records; {n_pts} unique patients)")
  ) |>
  cols_label(Characteristic = "Characteristic", `N (%) / Value` = "n (%) / Value") |>
  tab_style(
    style = list(cell_fill(color="#D9E1F2"), cell_text(weight="bold")),
    locations = cells_body(rows = section_rows)
  ) |>
  tab_source_note("SD = standard deviation. IQR = interquartile range. Age from column AP (days) ÷ 365.25. Age and sex N exclude missing values.") |>
  tab_options(table.width = pct(70), column_labels.font.weight = "bold")

gt_t1
```

---

# Table 2 — Test Volume, Validity, and Results

```{r table2-calcs}
#| label: table2-calcs

n_pos_all     <- sum(df_valid$is_pos, na.rm = TRUE)
n_neg_all     <- sum(df_valid$is_neg, na.rm = TRUE)
pct_pos_all   <- n_pos_all / n_valid * 100
pct_neg_all   <- n_neg_all / n_valid * 100
pct_qns       <- n_qns_n   / n_total * 100
pct_can       <- n_can     / n_total * 100

assay_summary <- df_valid |>
  mutate(assay_grp = case_when(
    str_detect(assay_std, "PIK3CA|PIK") ~ "PIK3CA Multiplex",
    str_detect(assay_std, "TEK")        ~ "TEK",
    str_detect(assay_std, "BRAF")       ~ "BRAF",
    TRUE                                ~ "Other"
  )) |>
  group_by(assay_grp) |>
  summarise(n=n(), n_pos=sum(is_pos), n_neg=sum(is_neg), .groups="drop") |>
  mutate(pct_pos = n_pos/n*100, pct_neg = n_neg/n*100)

spec_summary <- df_valid |>
  filter(!is.na(specimen_source_std)) |>
  group_by(specimen_source_std) |>
  summarise(n=n(), n_pos=sum(is_pos), n_neg=sum(is_neg), .groups="drop") |>
  mutate(pct_pos = n_pos/n*100, pct_neg = n_neg/n*100)

type_summary <- df_valid |>
  filter(!is.na(specimen_type_std)) |>
  group_by(specimen_type_std) |>
  summarise(n=n(), n_pos=sum(is_pos), n_neg=sum(is_neg), .groups="drop") |>
  mutate(pct_pos = n_pos/n*100, pct_neg = n_neg/n*100)

enc_both <- df_valid |>
  group_by(patient_id, collection_date) |>
  summarise(types = n_distinct(specimen_source_std), .groups="drop") |>
  filter(types > 1) |> nrow()

df_valid2 <- df_valid |>
  mutate(spec_grp = case_when(
    specimen_source_std == "Blood"           ~ "Blood only",
    specimen_source_std == "Lymphatic Fluid" ~ "Lymphatic fluid only",
    TRUE                                     ~ "Other"
  ))

n_blood_only <- sum(df_valid2$spec_grp == "Blood only",           na.rm = TRUE)
n_lymph_only <- sum(df_valid2$spec_grp == "Lymphatic fluid only", na.rm = TRUE)
n_both       <- enc_both
pct_blood    <- n_blood_only / n_valid * 100
pct_lymph    <- n_lymph_only / n_valid * 100
pct_both     <- n_both       / n_valid * 100

# Fluid volume uses df_patients (one entry per unique patient)
# Fluid volume: deduplicate df_valid by patient_id, exclude zeros and blanks
df_vol  <- df_valid |>
  filter(!is.na(vol_val) & vol_val > 0) |>
  group_by(patient_id) |>
  slice_min(row_number(), n = 1, with_ties = FALSE) |>
  ungroup()
vol_n   <- nrow(df_vol)
med_vol <- median(df_vol$vol_val, na.rm = TRUE)

# DNA and droplet metrics — all valid tests (test-level, no deduplication)
conc_n    <- sum(!is.na(df_valid$dna_conc_val))
used_n    <- sum(!is.na(df_valid$dna_used_val))
med_conc  <- median(df_valid$dna_conc_val, na.rm = TRUE)
mean_used <- mean(df_valid$dna_used_val,   na.rm = TRUE)
sd_used   <- sd(df_valid$dna_used_val,     na.rm = TRUE)
mean_ex21 <- mean(df_valid$ex21_val,       na.rm = TRUE)
sd_ex21   <- sd(df_valid$ex21_val,         na.rm = TRUE)
mean_ex10 <- mean(df_valid$ex10_val,       na.rm = TRUE)
sd_ex10   <- sd(df_valid$ex10_val,         na.rm = TRUE)

# Fluid volume pos/neg breakdown — unique sample_id_for_di_dataset, exclude zeros/blanks
df_vol_pos <- df_valid |> filter(is_pos, !is.na(vol_val), vol_val > 0) |>
              distinct(sample_id_for_di_dataset, .keep_all = TRUE)
df_vol_neg <- df_valid |> filter(is_neg, !is.na(vol_val), vol_val > 0) |>
              distinct(sample_id_for_di_dataset, .keep_all = TRUE)
med_vol_pos <- median(df_vol_pos$vol_val, na.rm = TRUE)
med_vol_neg <- median(df_vol_neg$vol_val, na.rm = TRUE)
vol_pos_n   <- nrow(df_vol_pos)
vol_neg_n   <- nrow(df_vol_neg)

# DNA/EXON pos/neg — all valid tests by result
med_conc_pos  <- median(df_valid$dna_conc_val[df_valid$is_pos], na.rm = TRUE)
med_conc_neg  <- median(df_valid$dna_conc_val[df_valid$is_neg], na.rm = TRUE)
mean_used_pos <- mean(df_valid$dna_used_val[df_valid$is_pos],   na.rm = TRUE)
sd_used_pos   <- sd(df_valid$dna_used_val[df_valid$is_pos],     na.rm = TRUE)
mean_used_neg <- mean(df_valid$dna_used_val[df_valid$is_neg],   na.rm = TRUE)
sd_used_neg   <- sd(df_valid$dna_used_val[df_valid$is_neg],     na.rm = TRUE)
mean_ex21_pos <- mean(df_valid$ex21_val[df_valid$is_pos],       na.rm = TRUE)
sd_ex21_pos   <- sd(df_valid$ex21_val[df_valid$is_pos],         na.rm = TRUE)
mean_ex21_neg <- mean(df_valid$ex21_val[df_valid$is_neg],       na.rm = TRUE)
sd_ex21_neg   <- sd(df_valid$ex21_val[df_valid$is_neg],         na.rm = TRUE)
mean_ex10_pos <- mean(df_valid$ex10_val[df_valid$is_pos],       na.rm = TRUE)
sd_ex10_pos   <- sd(df_valid$ex10_val[df_valid$is_pos],         na.rm = TRUE)
mean_ex10_neg <- mean(df_valid$ex10_val[df_valid$is_neg],       na.rm = TRUE)
sd_ex10_neg   <- sd(df_valid$ex10_val[df_valid$is_neg],         na.rm = TRUE)

ga <- function(name) {
  row <- assay_summary |> filter(assay_grp == name)
  if (nrow(row)==0) return(c(paste0("  ",name), "—","—","—"))
  c(paste0("  ",name),
    as.character(row$n),
    cell2(row$n_pos, row$pct_pos),
    cell2(row$n_neg, row$pct_neg))
}

gtype <- function(name, label) {
  row <- type_summary |> filter(specimen_type_std == name)
  if (nrow(row)==0) return(c(label,"—","—","—"))
  c(label,
    as.character(row$n),
    cell2(row$n_pos, row$pct_pos),
    cell2(row$n_neg, row$pct_neg))
}

gs <- function(name, label) {
  row <- spec_summary |> filter(specimen_source_std == name)
  if (nrow(row)==0) return(c(label,"—","—","—"))
  c(label,
    as.character(row$n),
    cell2(row$n_pos, row$pct_pos),
    cell2(row$n_neg, row$pct_neg))
}

t2_rows <- list(
  c("Test Validity","","",""),
  c("  Total tests ordered",  as.character(n_total),"",""),
  c("  Analyzable (valid)", fmt_np(n_valid, n_valid/n_records*100),
    cell2(n_pos_all,pct_pos_all), cell2(n_neg_all,pct_neg_all)),
  c("  QNS", fmt_np(n_qns_n, n_qns_n/n_records*100), "", ""),
  c("  Cancelled", fmt_np(n_can, n_can/n_records*100), "", ""),
  c("Results by Assay (valid tests only)","","",""),
  ga("PIK3CA Multiplex"),
  ga("TEK"),
  ga("BRAF"),
  c("Specimen Type (valid tests only)","","",""),
  gtype("Blood",        "  Blood"),
  gtype("Cyst Fluid",   "  Cyst Fluid"),
  gtype("Other", "  FFPE Scrolls"),
  c("Specimen Source (valid tests only)","","",""),
  gs("Blood",           "  Blood"),
  gs("Lymphatic Fluid", "  Lymphatic Fluid"),
  gs("Other",           "  Other"),
  c("Specimen Collection Metrics","","",""),
  c(sprintf("  Fluid volume, mL — median (N=%d patients)", vol_n),
    sprintf("%.1f", med_vol),
    sprintf("%.1f (N=%d)", med_vol_pos, vol_pos_n),
    sprintf("%.1f (N=%d)", med_vol_neg, vol_neg_n)),
  c(sprintf("  DNA concentration, ng/µL — median (N=%d)", conc_n),
    sprintf("%.1f", med_conc),
    sprintf("%.1f", med_conc_pos),
    sprintf("%.1f", med_conc_neg)),
  c(sprintf("  DNA used in assay, ng — mean (SD) (N=%d)", used_n),
    fmt_ms(mean_used, sd_used),
    fmt_ms(mean_used_pos, sd_used_pos),
    fmt_ms(mean_used_neg, sd_used_neg)),
  c(sprintf("  EXON 21 WT droplets — mean (SD) (N=%d)", sum(!is.na(df_valid$ex21_val))),
    fmt_ms(mean_ex21, sd_ex21),
    fmt_ms(mean_ex21_pos, sd_ex21_pos),
    fmt_ms(mean_ex21_neg, sd_ex21_neg)),
  c(sprintf("  EXON 10 WT droplets — mean (SD) (N=%d)", sum(!is.na(df_valid$ex10_val))),
    fmt_ms(mean_ex10, sd_ex10),
    fmt_ms(mean_ex10_pos, sd_ex10_pos),
    fmt_ms(mean_ex10_neg, sd_ex10_neg))
)

t2_df <- bind_rows(lapply(t2_rows, function(r)
  tibble(Category=r[1], `Total n`=r[2],
         `Positive n (%)`=r[3], `Negative n (%)`=r[4])))

section2_rows <- which(t2_df$Category %in% c(
  "Test Validity","Results by Assay (valid tests only)",
  "Specimen Type (valid tests only)",
  "Specimen Source (valid tests only)","Specimen Collection Metrics"))

gt_t2 <- t2_df |>
  gt() |>
  tab_header(
    title    = "Table 2. Test Volume, Validity, and Results",
    subtitle = glue("N = {n_total} tests")
  ) |>
  cols_label(
    Category       = "Category",
    `Total n`      = "n / Value",
    `Positive n (%)` = "Positive n (%)",
    `Negative n (%)` = "Negative n (%)"
  ) |>
  tab_style(
    style = list(cell_fill(color="#D9E1F2"), cell_text(weight="bold")),
    locations = cells_body(rows = section2_rows)
  ) |>
  tab_source_note("QNS = quantity not sufficient. Blank result entries excluded from all counts. NEG* treated as Negative. Fluid volume reported at patient level (unique patient ID); DNA and droplet metrics at test level (all valid tests). Positive/negative fluid volume uses unique sample IDs.") |>
  tab_options(table.width = pct(80), column_labels.font.weight = "bold")

gt_t2
```

---

# Table 3 — Assay Performance Metrics

```{r table3-calcs}
#| label: table3-calcs

# PIK3CA variant sub-rows (GNAQ excluded per protocol; TEK/BRAF renamed inline)
pik3ca_variants <- list(
  list(label = "  p.E542K",  pattern = "E542K"),
  list(label = "  p.E545K",  pattern = "E545K"),
  list(label = "  p.H1047R", pattern = "H1047R")
)

assay_list <- list(
  list(l="PIK3CA Multiplex", p="PIK3CA|PIK", ex=TRUE,
       pos_label="Positive",           neg_label="Negative"),
  list(l="TEK",              p="TEK",        ex=FALSE,
       pos_label="Positive (p.L914F)", neg_label="Negative"),
  list(l="BRAF",             p="BRAF",       ex=FALSE,
       pos_label="Positive (p.V600E)", neg_label="Negative")
)

# BUG FIX: specimen type counts computed per result group (not whole-assay totals)
spec_counts_r <- function(s) {
  n_r  <- nrow(s)
  n_bl <- sum(s$specimen_type_std == "Blood",      na.rm = TRUE)
  n_ly <- sum(s$specimen_type_std == "Cyst Fluid", na.rm = TRUE)
  list(
    bl = fmt_np(n_bl, if (n_r > 0) n_bl / n_r * 100 else 0),
    ly = fmt_np(n_ly, if (n_r > 0) n_ly / n_r * 100 else 0)
  )
}

t3_rows <- list()

for (ai in assay_list) {
  sub   <- df_valid |> filter(str_detect(assay_std, ai$p))
  nn    <- nrow(sub)
  s_pos <- filter(sub, is_pos)
  s_neg <- filter(sub, is_neg)

  # --- Assay header row ---
  t3_rows[[length(t3_rows)+1]] <- tibble(
    Assay                                  = ai$l,
    Result                                 = "",
    `n (%)`                                = "",
    `Fluid Vol (mL) Median [IQR]`          = "",
    `DNA Conc (ng/µL) Median [IQR]`        = "",
    `DNA Used (ng) Median [IQR]`           = "",
    `EXON 21 WT Droplets Median [IQR]`     = "",
    `EXON 10 WT Droplets Median [IQR]`     = "",
    `Median Single Probe Avg Droplets`     = "",
    `Specimen Type: Blood n (%)`           = "",
    `Specimen Type: Cyst Fluid n (%)`      = ""
  )

  # --- Positive row ---
  n_pos   <- nrow(s_pos)
  pct_pos <- if (nn > 0) n_pos / nn * 100 else 0
  s_vol_p <- s_pos |> filter(!is.na(vol_val) & vol_val > 0) |>
             distinct(sample_id_for_di_dataset, .keep_all = TRUE)
  sp_p    <- spec_counts_r(s_pos)

  t3_rows[[length(t3_rows)+1]] <- tibble(
    Assay                              = "",
    Result                             = ai$pos_label,
    `n (%)`                            = fmt_np(n_pos, pct_pos),
    `Fluid Vol (mL) Median [IQR]`      = iqr_s(s_vol_p$vol_val),
    `DNA Conc (ng/µL) Median [IQR]`    = iqr_s(s_pos$dna_conc_val),
    `DNA Used (ng) Median [IQR]`       = iqr_s(s_pos$dna_used_val),
    `EXON 21 WT Droplets Median [IQR]` = if (ai$ex) iqr_s(s_pos$ex21_val)         else "N/A",
    `EXON 10 WT Droplets Median [IQR]` = if (ai$ex) iqr_s(s_pos$ex10_val)         else "N/A",
    `Median Single Probe Avg Droplets` = if (!ai$ex) iqr_s(s_pos$single_probe_val) else "N/A",
    `Specimen Type: Blood n (%)`       = sp_p$bl,
    `Specimen Type: Cyst Fluid n (%)`  = sp_p$ly
  )

  # --- PIK3CA variant sub-rows ---
  if (ai$l == "PIK3CA Multiplex") {
    for (v in pik3ca_variants) {
      sv      <- s_pos |> filter(str_detect(variant_val, v$pattern))
      n_sv    <- nrow(sv)
      pct_sv  <- if (n_pos > 0) n_sv / n_pos * 100 else 0
      s_vol_v <- sv |> filter(!is.na(vol_val) & vol_val > 0) |>
                 distinct(sample_id_for_di_dataset, .keep_all = TRUE)
      sp_v    <- spec_counts_r(sv)
      t3_rows[[length(t3_rows)+1]] <- tibble(
        Assay                              = "",
        Result                             = v$label,
        `n (%)`                            = fmt_np(n_sv, pct_sv),
        `Fluid Vol (mL) Median [IQR]`      = iqr_s(s_vol_v$vol_val),
        `DNA Conc (ng/µL) Median [IQR]`    = iqr_s(sv$dna_conc_val),
        `DNA Used (ng) Median [IQR]`       = iqr_s(sv$dna_used_val),
        `EXON 21 WT Droplets Median [IQR]` = iqr_s(sv$ex21_val),
        `EXON 10 WT Droplets Median [IQR]` = iqr_s(sv$ex10_val),
        `Median Single Probe Avg Droplets` = "N/A",
        `Specimen Type: Blood n (%)`       = sp_v$bl,
        `Specimen Type: Cyst Fluid n (%)`  = sp_v$ly
      )
    }
  }

  # --- Negative row ---
  n_neg   <- nrow(s_neg)
  pct_neg <- if (nn > 0) n_neg / nn * 100 else 0
  s_vol_n <- s_neg |> filter(!is.na(vol_val) & vol_val > 0) |>
             distinct(sample_id_for_di_dataset, .keep_all = TRUE)
  sp_n    <- spec_counts_r(s_neg)

  t3_rows[[length(t3_rows)+1]] <- tibble(
    Assay                              = "",
    Result                             = ai$neg_label,
    `n (%)`                            = fmt_np(n_neg, pct_neg),
    `Fluid Vol (mL) Median [IQR]`      = iqr_s(s_vol_n$vol_val),
    `DNA Conc (ng/µL) Median [IQR]`    = iqr_s(s_neg$dna_conc_val),
    `DNA Used (ng) Median [IQR]`       = iqr_s(s_neg$dna_used_val),
    `EXON 21 WT Droplets Median [IQR]` = if (ai$ex) iqr_s(s_neg$ex21_val)         else "N/A",
    `EXON 10 WT Droplets Median [IQR]` = if (ai$ex) iqr_s(s_neg$ex10_val)         else "N/A",
    `Median Single Probe Avg Droplets` = if (!ai$ex) iqr_s(s_neg$single_probe_val) else "N/A",
    `Specimen Type: Blood n (%)`       = sp_n$bl,
    `Specimen Type: Cyst Fluid n (%)`  = sp_n$ly
  )

  # --- p-value row (Mann-Whitney U / Fisher's exact, Positive vs Negative) ---
  p_vol  <- fmt_p(safe_wilcox(s_vol_p$vol_val, s_vol_n$vol_val))
  p_conc <- fmt_p(safe_wilcox(s_pos$dna_conc_val,  s_neg$dna_conc_val))
  p_used <- fmt_p(safe_wilcox(s_pos$dna_used_val,  s_neg$dna_used_val))
  p_ex21 <- if (ai$ex)  fmt_p(safe_wilcox(s_pos$ex21_val, s_neg$ex21_val)) else "N/A"
  p_ex10 <- if (ai$ex)  fmt_p(safe_wilcox(s_pos$ex10_val, s_neg$ex10_val)) else "N/A"
  p_sprb <- if (!ai$ex) fmt_p(safe_wilcox(s_pos$single_probe_val, s_neg$single_probe_val)) else "N/A"
  sp_sub <- sub |> filter(specimen_type_std %in% c("Blood", "Cyst Fluid"))
  p_spec <- if (nrow(sp_sub) >= 2)
              fmt_p(safe_fisher(sp_sub$specimen_type_std, sp_sub$is_pos))
            else "--"

  t3_rows[[length(t3_rows)+1]] <- tibble(
    Assay                              = "",
    Result                             = "p-value",
    `n (%)`                            = "",
    `Fluid Vol (mL) Median [IQR]`      = p_vol,
    `DNA Conc (ng/µL) Median [IQR]`    = p_conc,
    `DNA Used (ng) Median [IQR]`       = p_used,
    `EXON 21 WT Droplets Median [IQR]` = p_ex21,
    `EXON 10 WT Droplets Median [IQR]` = p_ex10,
    `Median Single Probe Avg Droplets` = p_sprb,
    `Specimen Type: Blood n (%)`       = p_spec,
    `Specimen Type: Cyst Fluid n (%)`  = ""
  )
}

table3_df    <- bind_rows(t3_rows)
```

```{r table3-display}
#| label: table3-display

section3_rows <- which(table3_df$Assay != "")
pval3_rows    <- which(table3_df$Result == "p-value")
var3_rows     <- which(grepl("^  p\\.", table3_df$Result))

gt_t3 <- table3_df |>
  gt() |>
  tab_header(
    title    = "Table 3. Assay Performance Metrics by Result",
    subtitle = "Valid tests only"
  ) |>
  tab_style(
    style = list(cell_fill(color = "#D9E1F2"), cell_text(weight = "bold")),
    locations = cells_body(rows = section3_rows)
  ) |>
  tab_style(
    style = list(cell_text(style = "italic", color = "#595959")),
    locations = cells_body(rows = pval3_rows)
  ) |>
  tab_style(
    style = cell_text(indent = px(16)),
    locations = cells_body(columns = Result, rows = var3_rows)
  ) |>
  tab_source_note(paste(
    "IQR = interquartile range. All continuous metrics reported as Median [IQR].",
    "EXON 21/10 WT droplet counts: N/A for TEK and BRAF.",
    "Median single probe avg droplets: N/A for PIK3CA Multiplex.",
    "Specimen type breakdown reflects tests within each result group.",
    "p-values: Mann-Whitney U test for continuous variables;",
    "Fisher's exact test for specimen type (Blood vs Cyst Fluid).",
    "PIK3CA variant sub-row percentages are of PIK3CA Positive tests."
  )) |>
  tab_options(table.width = pct(100), column_labels.font.weight = "bold")

gt_t3
```

---

# Table 4 — Clinical Impact of Test Results

```{r table4-calcs}
#| label: table4-calcs

# Restrict to valid tests with POS or NEG result
df_t4 <- df_valid |> filter(is_pos | is_neg)

# N for Table 4 = total valid tests (pos or neg) with non-blank mgmt_change
# n (%) in each row = count of Y / N_total
all_mgmt   <- toupper(trimws(df_t4$mgmt_change))
n_t4_total <- sum(!is.na(all_mgmt) & all_mgmt != "" & all_mgmt != "NA")

# Helper: count Y (or specified values) across ALL valid tests, % out of n_t4_total
count_yn <- function(col, values_to_count = "Y") {
  vals      <- toupper(trimws(col))
  non_blank <- vals[!is.na(vals) & vals != "" & vals != "NA"]
  n_yes     <- sum(non_blank %in% values_to_count)
  pct       <- if (n_t4_total > 0) n_yes / n_t4_total * 100 else 0
  list(n = n_yes, fmt = fmt_np(n_yes, pct))
}

# Any management change — Y only
mgmt_res <- count_yn(df_t4$mgmt_change)

# Targeted therapy initiated — Y only
ti_res   <- count_yn(df_t4$therapy_initiated)

# Therapy modified — Y only
tm_res   <- count_yn(df_t4$therapy_modified)

# Procedure changed — Y only
pc_res   <- count_yn(df_t4$procedure_changed)

# No management change — N only
no_res   <- count_yn(df_t4$mgmt_change, values_to_count = "N")

# Mgmt unavailable — Y only
un_res   <- count_yn(df_t4$mgmt_unavailable)

# Additional testing — Y or VAN
at_res   <- count_yn(df_t4$additional_testing, values_to_count = c("Y","VAN"))

t4 <- tibble(
  `Clinical Action / Outcome` = c(
    "  Any documented management change",
    "    Targeted therapy initiated (e.g., alpelisib, sirolimus)",
    "    Existing therapy modified or dose-adjusted",
    "    Surgical or procedural plan changed",
    "  No documented management change",
    "  Management documentation unavailable / not reported",
    "Patients who underwent additional testing"
  ),
  `n (%)` = c(
    mgmt_res$fmt, ti_res$fmt, tm_res$fmt, pc_res$fmt,
    no_res$fmt, un_res$fmt, at_res$fmt
  )
)

section4_rows <- integer(0)  # no highlighted rows in Table 4
```

```{r table4-display}
#| label: table4-display

gt_t4 <- t4 |>
  gt() |>
  tab_header(title = glue("Table 4. Clinical Impact of Test Results (Total patients assessed, N = {n_t4_total})")) |>
  tab_style(
    style = list(cell_fill(color="#D9E1F2"), cell_text(weight="bold")),
    locations = cells_body(rows = section4_rows)
  ) |>
  tab_source_note("Clinical actions were assessed in internally managed patients with valid (non-QNS, non-cancelled) results, as complete clinical documentation was available only for this subgroup. Y and VAN counted as positive for additional testing. Blank entries excluded from denominators. Targeted therapies listed are examples; actual agents confirmed by chart review.") |>
  tab_options(table.width = pct(70), column_labels.font.weight = "bold")

gt_t4
```

---

# Data Accounting

```{r data-accounting-summary}
#| label: data-accounting-summary

tibble(
  Item  = c(
    "Total rows in raw file",
    "Spacer rows excluded (no Patient ID)",
    "Records retained for analysis",
    "Unique patients"
  ),
  Count = c(n_raw, n_spacer, n_records, n_pts)
) |>
  gt() |>
  tab_header(title = "Data Accounting Summary") |>
  tab_options(table.width = pct(40), column_labels.font.weight = "bold")
```

> **Missing data policy:** All records with a valid Patient ID are retained (N = `r n_records`). For each analysis, only rows with valid (non-QNS, non-cancelled, non-blank) results are included. Blank result entries (N = `r n_blanks`) are excluded from all counts. Age and sex N reflect records with non-missing values only.

<!-- HTML stays local. Run spotseq_run_analysis_v2.R to export Excel to Google Drive. -->
