---
title: "SpotSeq Dataset Analysis"
author: "SpotSeq Analysis Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(readxl)
library(janitor)
library(openxlsx)
library(googledrive)
library(knitr)
library(kableExtra)
library(glue)

data_path       <- Sys.getenv("DATA_PATH")
drive_folder_id <- Sys.getenv("DRIVE_FOLDER_ID")

stopifnot(
  "DATA_PATH not set in .Renviron"       = nchar(data_path) > 0,
  "DRIVE_FOLDER_ID not set in .Renviron" = nchar(drive_folder_id) > 0
)
```

```{r load-data}
#| label: load-data

raw <- read_excel(
  file.path(data_path, "SpotSeq_Project_Dataset_deidentified_2.1.xlsx")
)
df_raw    <- raw |> clean_names()
df        <- df_raw |> filter(!is.na(patient_id) & patient_id != "")
n_raw     <- nrow(raw)
n_spacer  <- n_raw - nrow(df)
n_records <- nrow(df)
n_pts     <- n_distinct(df$patient_id)
```

```{r specimen-standardise}
#| label: specimen-standardise

blood_vals <- toupper(c("Blood","Blood, venous","Lesional blood"))
lymph_vals <- toupper(c(
  "Lymphatic fluid","Lymphatic fluid, Site A","Lymphatic fluid, Site B",
  "Lymphatic fluid from left groin","Lymphatic cyst fluid",
  "Lymphatic cyst fluid from lymph node","Cyst fluid",
  "Lymphatic Malformation","Lymphatic malformation",
  "Chest, Lymphatic malformation"
))
other_vals <- toupper(c(
  "Aspirated Cyst Fluid","aspirated cyst fluid",
  "Aspirate from the cyst in the abdomen",
  "Abdomen, Lymphatic Cyst Fluid","Ankle, Left","Neck",
  "Tracheocutaneous fistula","Peritoneal fluid"
))

df <- df |>
  mutate(
    specimen_source_raw = toupper(trimws(specimen_source)),
    specimen_source_std = case_when(
      specimen_source_raw %in% blood_vals ~ "Blood",
      specimen_source_raw %in% lymph_vals ~ "Lymphatic Fluid",
      specimen_source_raw %in% other_vals ~ "Other",
      TRUE                                ~ NA_character_
    )
  )
```

```{r classify-results}
#| label: classify-results

df <- df |>
  mutate(
    result_clean = toupper(trimws(result_pos_neg)),
    status_clean = toupper(trimws(status)),
    is_qns       = status_clean == "QNS" |
                   result_clean %in% c("QNS","QNS FOR REPEAT"),
    is_cancelled = status_clean %in% c("CANCELLED","CANCELED") |
                   result_clean %in% c("CANCELLED","CANCELED"),
    is_valid     = !is_qns & !is_cancelled &
                   status_clean == "COMPLETE" & !is.na(result_clean),
    is_pos       = is_valid & grepl("^POS", result_clean),
    is_neg       = is_valid & grepl("^NEG", result_clean),
    assay_std    = toupper(trimws(assay_ordered)),
    age_val      = suppressWarnings(
                     as.numeric(age_at_date_of_collection_in_years)),
    sex_val      = assigned_sex_at_birth,
    internal_val = toupper(trimws(internal_patient_y_n)),
    vol_val      = suppressWarnings({
      v <- trimws(fluid_volume_extracted)
      ifelse(grepl("/", v),
             sapply(strsplit(v, "/"), function(x)
               sum(as.numeric(trimws(x)), na.rm = TRUE)),
             as.numeric(v))
    }),
    dna_conc_val = suppressWarnings(as.numeric(
      dna_stock_ng_ul_blood_0_185ng_ul_reject_cyst_fluid_0_37ng_ul_reject)),
    dna_used_val = suppressWarnings(as.numeric(total_ng_dna_used_in_assay)),
    drop_val     = suppressWarnings(
                     as.numeric(exon_21_wt_1500_avg_number_of_droplets)),
    ex21_val     = suppressWarnings(
                     as.numeric(exon_21_wt_1500_avg_number_of_droplets)),
    ex10_val     = suppressWarnings(
                     as.numeric(exon_10_wt_1500_avg_number_droplets))
  )

df_valid  <- df |> filter(is_valid)
df_qns    <- df |> filter(is_qns)
df_cancel <- df |> filter(is_cancelled)
```

```{r helpers}
#| label: helpers

fmt_np  <- function(n, p) sprintf("%d (%.1f%%)", as.integer(n), p)
fmt_ms  <- function(m, s) sprintf("%.1f (%.1f)", m, s)
cell2   <- function(n, p) sprintf("%d (%.1f%%)", as.integer(n), p)

iqr_s <- function(x) {
  if (sum(!is.na(x)) == 0) return("\u2014")
  q <- quantile(x, c(.25, .75), na.rm = TRUE)
  sprintf("%.1f (%.1f\u2013%.1f)", median(x, na.rm=TRUE), q[1], q[2])
}
msd_s <- function(x) {
  if (sum(!is.na(x)) == 0) return("\u2014")
  sprintf("%.1f (%.1f)", mean(x, na.rm=TRUE), sd(x, na.rm=TRUE))
}
spec_bk <- function(s, nn) {
  nb <- sum(s$specimen_source_std == "Blood",           na.rm=TRUE)
  nl <- sum(s$specimen_source_std == "Lymphatic Fluid", na.rm=TRUE)
  bt <- s |>
    group_by(patient_id, collection_date) |>
    summarise(t = n_distinct(specimen_source_std), .groups="drop") |>
    filter(t > 1) |> nrow()
  list(
    blood = if (nn > 0) cell2(nb, nb/nn*100) else "\u2014",
    lymph = if (nn > 0) cell2(nl, nl/nn*100) else "\u2014",
    both  = if (nn > 0) cell2(bt, bt/nn*100) else "\u2014"
  )
}
```

```{r table1-calcs}
#| label: table1-calcs

age_n      <- sum(!is.na(df$age_val))
age_mean   <- mean(df$age_val, na.rm=TRUE)
age_sd     <- sd(df$age_val,   na.rm=TRUE)
n_peds     <- sum(df$age_val <  18, na.rm=TRUE)
n_adult    <- sum(df$age_val >= 18, na.rm=TRUE)
pct_peds   <- n_peds  / age_n * 100
pct_adult  <- n_adult / age_n * 100
sex_clean  <- toupper(trimws(df$sex_val))
n_female   <- sum(sex_clean == "F", na.rm=TRUE)
sex_n      <- sum(!is.na(df$sex_val))
pct_female <- n_female / sex_n * 100
n_tests    <- nrow(df)
n_internal <- sum(df$internal_val == "Y", na.rm=TRUE)
pct_int    <- n_internal / n_tests * 100
```

```{r table2-calcs}
#| label: table2-calcs

n_total     <- nrow(df)
n_valid     <- nrow(df_valid)
n_qns_n     <- nrow(df_qns)
n_can       <- nrow(df_cancel)
pct_qns     <- n_qns_n / n_total * 100
pct_can     <- n_can   / n_total * 100
n_pos_all   <- sum(df_valid$is_pos)
n_neg_all   <- sum(df_valid$is_neg)
pct_pos_all <- n_pos_all / n_valid * 100
pct_neg_all <- n_neg_all / n_valid * 100

assay_summary <- df_valid |>
  mutate(assay_grp = case_when(
    str_detect(assay_std, "PIK3CA|PIK") ~ "PIK3CA Multiplex",
    str_detect(assay_std, "TEK")        ~ "TEK",
    str_detect(assay_std, "BRAF")       ~ "BRAF",
    TRUE                                ~ "Other"
  )) |>
  group_by(assay_grp) |>
  summarise(n=n(), n_pos=sum(is_pos), n_neg=sum(is_neg), .groups="drop") |>
  mutate(pct_pos=n_pos/n*100, pct_neg=n_neg/n*100)

spec_summary <- df_valid |>
  filter(!is.na(specimen_source_std)) |>
  group_by(specimen_source_std) |>
  summarise(n=n(), n_pos=sum(is_pos), n_neg=sum(is_neg), .groups="drop") |>
  mutate(pct_pos=n_pos/n*100, pct_neg=n_neg/n*100)

enc_both <- df_valid |>
  group_by(patient_id, collection_date) |>
  summarise(types=n_distinct(specimen_source_std), .groups="drop") |>
  filter(types > 1) |> nrow()

df_valid2  <- df_valid |>
  mutate(spec_grp = case_when(
    specimen_source_std == "Blood"           ~ "Blood only",
    specimen_source_std == "Lymphatic Fluid" ~ "Lymphatic fluid only",
    TRUE                                     ~ "Other"
  ))
n_blood_only <- sum(df_valid2$spec_grp == "Blood only",           na.rm=TRUE)
n_lymph_only <- sum(df_valid2$spec_grp == "Lymphatic fluid only", na.rm=TRUE)
n_both       <- enc_both
pct_blood    <- n_blood_only / n_valid * 100
pct_lymph    <- n_lymph_only / n_valid * 100
pct_both     <- n_both       / n_valid * 100

vol_n    <- sum(!is.na(df_valid$vol_val))
conc_n   <- sum(!is.na(df_valid$dna_conc_val))
used_n   <- sum(!is.na(df_valid$dna_used_val))
drop_n   <- sum(!is.na(df_valid$drop_val))
med_vol  <- median(df_valid$vol_val,      na.rm=TRUE)
med_conc <- median(df_valid$dna_conc_val, na.rm=TRUE)
mean_used<- mean(df_valid$dna_used_val,   na.rm=TRUE)
mean_drop<- mean(df_valid$drop_val,       na.rm=TRUE)
```

```{r table3-calcs}
#| label: table3-calcs

t3_rows <- list()
for (ai in list(
  list(l="PIK3CA MULTIPLEX", p="PIK3CA|PIK", ex=TRUE),
  list(l="TEK",              p="TEK",         ex=FALSE),
  list(l="BRAF",             p="BRAF",        ex=FALSE)
)) {
  sub <- df_valid |> filter(str_detect(assay_std, ai$p))
  for (res in c("Pos","Neg")) {
    s  <- if (res=="Pos") filter(sub,is_pos) else filter(sub,is_neg)
    nn <- nrow(s); sp <- spec_bk(s, nn)
    t3_rows[[length(t3_rows)+1]] <- tibble(
      Assay                       = if (res=="Pos") ai$l else "",
      Result                      = res,
      `Fluid Vol (µL)`            = iqr_s(s$vol_val),
      `DNA Conc (ng/µL)`          = iqr_s(s$dna_conc_val),
      `DNA Used (ng)`             = msd_s(s$dna_used_val),
      `Droplets/Well`             = msd_s(s$drop_val),
      `EXON 21 WT med(IQR)`      = if (ai$ex) iqr_s(s$ex21_val) else "N/A",
      `EXON 10 WT med(IQR)`      = if (ai$ex) iqr_s(s$ex10_val) else "N/A",
      `Blood Only n(%)`           = sp$blood,
      `Lymphatic Only n(%)`       = sp$lymph,
      `Both n(%)`                 = sp$both
    )
  }
}
table3_df <- bind_rows(t3_rows)
```

---

# Data Accounting

```{r data-accounting}
#| label: data-accounting

tibble(
  Item  = c("Total rows in raw file",
            "Spacer rows excluded (no Patient ID)",
            "Records retained for analysis",
            "Unique patients"),
  Count = c(n_raw, n_spacer, n_records, n_pts)
) |>
  kbl(caption = "Data Accounting Summary") |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE)
```

> **Missing data policy:** All records with a valid Patient ID are retained (N = `r n_records`). For each analysis, only rows with valid (non-NA) data for that specific variable are used; every result states the contributing N. Cancelled and QNS tests are counted separately and excluded from POS/NEG comparisons.

---

# Table 1 — Patient and Cohort Characteristics

```{r table1-display}
#| label: table1-display

t1 <- tibble(
  Characteristic = c(
    "Demographics",
    "  Total patients",
    "  Age, mean (SD), years",
    "  Age group",
    "    Pediatric (< 18 years)",
    "    Adult (\u2265 18 years)",
    "  Female sex, n (%)",
    "Testing Overview",
    "  Total tests ordered",
    "  Internally processed tests, n (%)"
  ),
  `n (%) / Value` = c(
    "",
    as.character(n_pts),
    glue("{fmt_ms(age_mean, age_sd)} (N = {age_n} of {n_records})"),
    "",
    glue("{fmt_np(n_peds,  pct_peds)}  (N = {age_n} of {n_records})"),
    glue("{fmt_np(n_adult, pct_adult)} (N = {age_n} of {n_records})"),
    glue("{fmt_np(n_female, pct_female)} (N = {sex_n} of {n_records})"),
    "",
    as.character(n_tests),
    fmt_np(n_internal, pct_int)
  )
)

section_rows <- which(t1$Characteristic %in%
                        c("Demographics","Testing Overview"))

t1 |>
  kbl(caption = glue(
    "Table 1. Patient and Cohort Characteristics ",
    "(N = {n_records} records; {n_pts} unique patients)"
  )) |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE) |>
  row_spec(section_rows, bold = TRUE, background = "#D9E1F2")
```

*SD = standard deviation. ddPCR = droplet digital polymerase chain reaction. Age and sex N exclude records with missing values.*

---

# Table 2 — Test Volume, Validity, and Results

```{r table2-display}
#| label: table2-display

ga <- function(name) {
  row <- assay_summary |> filter(assay_grp == name)
  if (nrow(row)==0) return(c(paste0("  ",name),"—","—","—","—"))
  c(paste0("  ",name), as.character(row$n),
    cell2(row$n_pos,row$pct_pos), cell2(row$n_neg,row$pct_neg),"—")
}
gs <- function(name, label) {
  row <- spec_summary |> filter(specimen_source_std == name)
  if (nrow(row)==0) return(c(label,"—","—","—","—"))
  c(label, as.character(row$n),
    cell2(row$n_pos,row$pct_pos), cell2(row$n_neg,row$pct_neg),"—")
}

t2_rows <- list(
  c("Test Validity","","","",""),
  c("  Total tests ordered",  as.character(n_total),"","",""),
  c("  Analyzable (valid)",   as.character(n_valid),
    cell2(n_pos_all,pct_pos_all), cell2(n_neg_all,pct_neg_all),"—"),
  c("  QNS",      as.character(n_qns_n),"—","—",cell2(n_qns_n,pct_qns)),
  c("  Cancelled", as.character(n_can), "—","—",cell2(n_can,pct_can)),
  c("Results by Assay (valid tests only)","","","",""),
  ga("PIK3CA Multiplex"), ga("TEK"), ga("BRAF"),
  c("Results by Specimen Type (valid tests only)","","","",""),
  c("  Blood only",
    as.character(n_blood_only), cell2(n_blood_only,pct_blood),"—","—"),
  c("  Lymphatic fluid only",
    as.character(n_lymph_only), cell2(n_lymph_only,pct_lymph),"—","—"),
  c("  Both blood & lymphatic fluid",
    as.character(n_both),       cell2(n_both,pct_both),        "—","—"),
  c("Lab QC Metrics (valid tests only)","","","",""),
  c(sprintf("  Median fluid volume (N=%d/%d)",   vol_n,  n_valid),
    sprintf("%.1f \u00b5L", med_vol),"","",""),
  c(sprintf("  Median DNA stock conc. (N=%d/%d)", conc_n, n_valid),
    sprintf("%.2f ng/\u00b5L", med_conc),"","",""),
  c(sprintf("  Mean total DNA used (N=%d/%d)",   used_n, n_valid),
    sprintf("%.1f ng", mean_used),"","",""),
  c(sprintf("  Mean droplets/well (N=%d/%d)",    drop_n, n_valid),
    sprintf("%.0f", mean_drop),"","",""),
  c("Results by Specimen Source Category (valid tests only)","","","",""),
  gs("Blood",          "  Blood (venous / lesional)"),
  gs("Lymphatic Fluid","  Lymphatic fluid"),
  gs("Other",          "  Other anatomic site")
)

t2_df <- do.call(rbind, t2_rows) |>
  as.data.frame() |>
  setNames(c("Category","Total n",
             "Positive n (%)","Negative n (%)","QNS/Cancelled n (%)"))

sec_rows_t2 <- which(t2_df$Category %in% c(
  "Test Validity",
  "Results by Assay (valid tests only)",
  "Results by Specimen Type (valid tests only)",
  "Lab QC Metrics (valid tests only)",
  "Results by Specimen Source Category (valid tests only)"
))

t2_df |>
  kbl(caption = glue(
    "Table 2. Test Volume, Validity, and Results ",
    "(N = {n_records} records)"
  )) |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = TRUE, font_size = 13) |>
  row_spec(sec_rows_t2, bold = TRUE, background = "#D9E1F2")
```

*QNS = quantity not sufficient. Percentages for QNS/cancelled are of total tests ordered; all other percentages are of valid (analyzable) tests.*

---

# Table 3 — Test Performance by Assay Type

```{r table3-display}
#| label: table3-display

table3_df |>
  kbl(caption = glue(
    "Table 3. Test Performance by Assay Type ",
    "(valid tests only; N = {n_valid} of {n_records} records)"
  )) |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = TRUE, font_size = 12) |>
  column_spec(1, bold = TRUE) |>
  add_header_above(c(
    " " = 2,
    "Laboratory Quality Control Metrics" = 6,
    "Result by Specimen Type" = 3
  ))
```

*IQR = interquartile range. EXON 21 and EXON 10 WT droplet metrics apply to PIK3CA Multiplex only (N/A for TEK and BRAF). Values reported as median (IQR) or mean (SD).*

---

# Export to Excel & Upload to Google Drive

```{r export-upload}
#| label: export-upload

source("spotseq_run_analysis.R")
```

---

# Session Info

```{r session-info}
#| label: session-info

sessionInfo()
```
