---
title: "SpotSeq Dataset Analysis"
author: "SpotSeq Analysis Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(readxl)
library(janitor)
library(openxlsx)
library(googledrive)
library(knitr)
library(kableExtra)
library(glue)

data_path      <- Sys.getenv("DATA_PATH")
drive_folder_id <- Sys.getenv("DRIVE_FOLDER_ID")

stopifnot(
  "DATA_PATH not set in .Renviron"      = nchar(data_path) > 0,
  "DRIVE_FOLDER_ID not set in .Renviron" = nchar(drive_folder_id) > 0
)
```

```{r load-data}
#| label: load-data

# Read all columns as text first to capture displayed cell values, not formulas
raw <- read_excel(
  file.path(data_path, "SpotSeq_Project_Dataset_deidentified_2.1.xlsx"),
  col_types = "text"
)

df_raw    <- raw |> clean_names()
df        <- df_raw |> filter(!is.na(patient_id) & patient_id != "")
n_raw     <- nrow(raw)
n_spacer  <- n_raw - nrow(df)
n_records <- nrow(df)
n_pts     <- n_distinct(df$patient_id)
```

```{r specimen-standardise}
#| label: specimen-standardise

blood_vals <- toupper(c("Blood","Blood, venous","Lesional blood"))
lymph_vals <- toupper(c(
  "Lymphatic fluid","Lymphatic fluid, Site A","Lymphatic fluid, Site B",
  "Lymphatic fluid from left groin","Lymphatic cyst fluid",
  "Lymphatic cyst fluid from lymph node","Cyst fluid",
  "Lymphatic Malformation","Lymphatic malformation",
  "Chest, Lymphatic malformation"
))
other_vals <- toupper(c(
  "Aspirated Cyst Fluid","aspirated cyst fluid",
  "Abdomen, Lymphatic Cyst Fluid","Ankle, Left","Neck",
  "Tracheocutaneous fistula","Peritoneal fluid"
))

df <- df |>
  mutate(
    specimen_source_raw = toupper(trimws(specimen_source)),
    specimen_source_std = case_when(
      specimen_source_raw %in% blood_vals ~ "Blood",
      specimen_source_raw %in% lymph_vals ~ "Lymphatic Fluid",
      specimen_source_raw %in% other_vals ~ "Other",
      TRUE                                ~ NA_character_
    )
  )
```

```{r classify-results}
#| label: classify-results

df <- df |>
  mutate(
    # Standardise result: trim, upper
    result_clean   = toupper(trimws(result_pos_neg)),
    status_clean   = toupper(trimws(status)),

    # Classification flags
    is_qns        = status_clean %in% c("QNS","QNS FOR REPEAT") |
                    result_clean %in% c("QNS","QNS FOR REPEAT"),
    is_cancelled  = status_clean %in% c("CANCELLED","CANCELED") |
                    result_clean %in% c("CANCELLED","CANCELED"),

    # Blank result entries — omit from analysis
    is_blank      = is.na(result_pos_neg) | trimws(result_pos_neg) == "",

    is_valid      = !is_qns & !is_cancelled & !is_blank &
                    status_clean == "COMPLETE" & !is.na(result_clean),

    # Positive / Negative: NEG* and any capitalisation variant treated as NEG
    is_pos        = is_valid & grepl("^POS", result_clean),
    is_neg        = is_valid & grepl("^NEG", result_clean),   # catches NEG, NEG*, Neg, neg, etc.

    assay_std     = toupper(trimws(assay_ordered)),

    # Age: use displayed value from column AP (age_at_date_of_collection_in_days)
    # Divide by 365.25 for decimal years — handles infants correctly
    age_days      = suppressWarnings(as.numeric(age_at_date_of_collection_in_days)),
    age_years     = age_days / 365.25,

    # Sex
    sex_val       = assigned_sex_at_birth,

    # Internal vs external patient (AQ)
    internal_val  = toupper(trimws(internal_patient_y_n)),

    # Volume / DNA / assay metrics — read from displayed values
    vol_val       = suppressWarnings(as.numeric(fluid_volume_extracted)),
    dna_conc_val  = suppressWarnings(as.numeric(
                      dna_stock_ng_ul_blood_0_185ng_ul_reject_cyst_fluid_0_37ng_ul_reject)),
    dna_used_val  = suppressWarnings(as.numeric(total_ng_dna_used_in_assay)),
    ex21_val      = suppressWarnings(as.numeric(exon_21_wt_1500_avg_number_of_droplets)),
    ex10_val      = suppressWarnings(as.numeric(exon_10_wt_1500_avg_number_droplets)),

    # Table 4 — Clinical impact columns (AR–AW), read as displayed values
    mgmt_change        = toupper(trimws(any_documented_management_change_in_management_post_testing)),
    therapy_initiated  = toupper(trimws(targeted_therapy_initiated_post_testing_e_g_alpelisib_sirolimus)),
    therapy_modified   = toupper(trimws(existing_therapy_modified_or_dose_adjusted_post_testing)),
    procedure_changed  = toupper(trimws(surgical_or_procedural_plan_changed_post_testing)),
    additional_testing = toupper(trimws(patient_underwent_additional_testing_based_on_spot_seq_results)),
    mgmt_unavailable   = toupper(trimws(management_documentation_post_testing_unavailable_not_reported))
  )

# Filtered subsets
df_valid    <- df |> filter(is_valid)
df_qns      <- df |> filter(is_qns)
df_cancel   <- df |> filter(is_cancelled)
# Omit blanks from all counts
df_analysed <- df |> filter(!is_blank)
```

```{r data-reconciliation}
#| label: data-reconciliation

# Explicit accounting of all records — surfaces any unclassified entries
n_total    <- nrow(df_analysed)
n_valid    <- nrow(df_valid)
n_qns_n    <- nrow(df_qns)
n_can      <- nrow(df_cancel)
n_blanks   <- sum(df$is_blank, na.rm = TRUE)
n_other    <- n_total - n_valid - n_qns_n - n_can   # should be 0

tibble(
  Category = c(
    "Total rows in raw file",
    "Spacer rows excluded (no Patient ID)",
    "Records retained (with Patient ID)",
    "  Blank result entries (omitted from analysis)",
    "  Records included in analysis",
    "    Valid (analyzable)",
    "    QNS",
    "    Cancelled",
    "    Unclassified (investigate if > 0)"
  ),
  Count = c(
    n_raw, n_spacer, n_records,
    n_blanks, n_total,
    n_valid, n_qns_n, n_can, n_other
  )
) |>
  kbl(caption = "Data Reconciliation — Record Accounting") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r helpers}
#| label: helpers

fmt_ms  <- function(m, s) sprintf("%.1f (%.1f)", m, s)
fmt_med <- function(med, lo, hi) sprintf("%.1f [%.1f–%.1f]", med, lo, hi)
fmt_np  <- function(n, p) sprintf("%d (%.1f%%)", as.integer(n), p)
cell2   <- function(n, p) sprintf("%d (%.1f%%)", as.integer(n), p)
iqr_s   <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return("—")
  sprintf("%.1f [%.1f–%.1f]", median(x), quantile(x,.25), quantile(x,.75))
}
msd_s   <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return("—")
  sprintf("%.1f (%.1f)", mean(x), sd(x))
}
spec_bk <- function(s, nn) {
  enc <- df_valid |>
    filter(specimen_source_std == s) |>
    group_by(patient_id, collection_date) |>
    summarise(types = n_distinct(specimen_source_std), .groups = "drop") |>
    filter(types > 1) |> nrow()
  bl   <- sum(df_valid$specimen_source_std == s & df_valid$is_pos, na.rm = TRUE)
  ly   <- sum(df_valid$specimen_source_std == s & df_valid$is_pos, na.rm = TRUE)
  both <- enc
  n_bl <- sum(df_valid$spec_grp == "Blood only",           na.rm = TRUE)
  n_ly <- sum(df_valid$spec_grp == "Lymphatic fluid only", na.rm = TRUE)
  list(
    blood  = fmt_np(n_bl, n_bl / nn * 100),
    lymph  = fmt_np(n_ly, n_ly / nn * 100),
    both   = fmt_np(both, both / nn * 100)
  )
}
```

```{r table1-calcs}
#| label: table1-calcs

# Age summary — using decimal years from AP column
age_n      <- sum(!is.na(df_valid$age_years))
age_mean   <- mean(df_valid$age_years,   na.rm = TRUE)
age_sd     <- sd(df_valid$age_years,     na.rm = TRUE)
age_med    <- median(df_valid$age_years, na.rm = TRUE)
age_q1     <- quantile(df_valid$age_years, 0.25, na.rm = TRUE)
age_q3     <- quantile(df_valid$age_years, 0.75, na.rm = TRUE)
age_min    <- min(df_valid$age_years,    na.rm = TRUE)
age_max    <- max(df_valid$age_years,    na.rm = TRUE)

# Pediatric / Adult split (< 18 years)
n_peds     <- sum(df_valid$age_years < 18,  na.rm = TRUE)
n_adult    <- sum(df_valid$age_years >= 18, na.rm = TRUE)
pct_peds   <- n_peds  / age_n * 100
pct_adult  <- n_adult / age_n * 100

# Sex
sex_clean  <- toupper(trimws(df_valid$sex_val))
n_female   <- sum(sex_clean == "F", na.rm = TRUE)
sex_n      <- sum(!is.na(df_valid$sex_val))
n_sex_n    <- sum(is.na(df_valid$sex_val))
pct_female <- n_female / sex_n * 100

# Internal vs External
internal_clean <- toupper(trimws(df_valid$internal_val))
n_internal     <- sum(internal_clean == "Y", na.rm = TRUE)
n_external     <- sum(internal_clean == "N", na.rm = TRUE)
n_tests        <- nrow(df_valid)
pct_int        <- n_internal / n_tests * 100
pct_ext        <- n_external / n_tests * 100
```

---

# Table 1 — Patient and Cohort Characteristics

```{r table1-display}
#| label: table1-display

t1 <- tibble(
  Characteristic = c(
    "Demographics",
    "  Age at collection, years",
    "    Mean (SD)",
    "    Median [IQR]",
    "    Range",
    "  Age group",
    "    Pediatric (< 18 years)",
    "    Adult (≥ 18 years)",
    "  Female sex, n (%)",
    "Patient Origin",
    "  Internally processed, n (%)",
    "  Externally referred, n (%)",
    "Testing Overview",
    "  Total tests ordered",
    "  Internally processed tests, n (%)"
  ),
  `N (%) / Value` = c(
    "",
    "",
    fmt_ms(age_mean, age_sd),
    fmt_med(age_med, age_q1, age_q3),
    sprintf("%.2f – %.1f", age_min, age_max),
    "",
    fmt_np(n_peds,  pct_peds),
    fmt_np(n_adult, pct_adult),
    glue("{fmt_np(n_female, pct_female)} (N = {sex_n} of {n_records})"),
    "",
    fmt_np(n_internal, pct_int),
    fmt_np(n_external, pct_ext),
    "",
    as.character(n_tests),
    fmt_np(n_internal, pct_int)
  )
)

section_rows <- which(t1$Characteristic %in%
                        c("Demographics","Patient Origin","Testing Overview"))

t1 |>
  kbl(caption = glue(
    "Table 1. Patient and Cohort Characteristics ",
    "(N = {n_records} records; {n_pts} unique patients)"
  )) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) |>
  row_spec(section_rows, bold = TRUE, background = "#D9E1F2")
```

*SD = standard deviation. IQR = interquartile range. Age calculated from displayed values in column AP (age in days) divided by 365.25. Age and sex N exclude records with missing values.*

---

# Table 2 — Test Volume, Validity, and Results

```{r table2-calcs}
#| label: table2-calcs

n_pos_all     <- sum(df_valid$is_pos, na.rm = TRUE)
n_neg_all     <- sum(df_valid$is_neg, na.rm = TRUE)
pct_pos_all   <- n_pos_all / n_valid * 100
pct_neg_all   <- n_neg_all / n_valid * 100
pct_qns       <- n_qns_n   / n_total * 100
pct_can       <- n_can     / n_total * 100

assay_summary <- df_valid |>
  mutate(assay_grp = case_when(
    str_detect(assay_std, "PIK3CA|PIK") ~ "PIK3CA Multiplex",
    str_detect(assay_std, "TEK")        ~ "TEK",
    str_detect(assay_std, "BRAF")       ~ "BRAF",
    TRUE                                ~ "Other"
  )) |>
  group_by(assay_grp) |>
  summarise(n=n(), n_pos=sum(is_pos), n_neg=sum(is_neg), .groups="drop") |>
  mutate(pct_pos = n_pos/n*100, pct_neg = n_neg/n*100)

spec_summary <- df_valid |>
  filter(!is.na(specimen_source_std)) |>
  group_by(specimen_source_std) |>
  summarise(n=n(), n_pos=sum(is_pos), n_neg=sum(is_neg), .groups="drop") |>
  mutate(pct_pos = n_pos/n*100, pct_neg = n_neg/n*100)

enc_both <- df_valid |>
  group_by(patient_id, collection_date) |>
  summarise(types = n_distinct(specimen_source_std), .groups="drop") |>
  filter(types > 1) |> nrow()

df_valid2 <- df_valid |>
  mutate(spec_grp = case_when(
    specimen_source_std == "Blood"           ~ "Blood only",
    specimen_source_std == "Lymphatic Fluid" ~ "Lymphatic fluid only",
    TRUE                                     ~ "Other"
  ))

n_blood_only <- sum(df_valid2$spec_grp == "Blood only",           na.rm = TRUE)
n_lymph_only <- sum(df_valid2$spec_grp == "Lymphatic fluid only", na.rm = TRUE)
n_both       <- enc_both
pct_blood    <- n_blood_only / n_valid * 100
pct_lymph    <- n_lymph_only / n_valid * 100
pct_both     <- n_both       / n_valid * 100

vol_n    <- sum(!is.na(df_valid$vol_val))
conc_n   <- sum(!is.na(df_valid$dna_conc_val))
used_n   <- sum(!is.na(df_valid$dna_used_val))
med_vol  <- median(df_valid$vol_val,      na.rm = TRUE)
med_conc <- median(df_valid$dna_conc_val, na.rm = TRUE)
mean_used <- mean(df_valid$dna_used_val,  na.rm = TRUE)
sd_used   <- sd(df_valid$dna_used_val,    na.rm = TRUE)
mean_ex21 <- mean(df_valid$ex21_val,      na.rm = TRUE)
sd_ex21   <- sd(df_valid$ex21_val,        na.rm = TRUE)

ga <- function(name) {
  row <- assay_summary |> filter(assay_grp == name)
  if (nrow(row)==0) return(c(paste0("  ",name), "—","—","—"))
  c(paste0("  ",name),
    as.character(row$n),
    cell2(row$n_pos, row$pct_pos),
    cell2(row$n_neg, row$pct_neg))
}

gs <- function(name, label) {
  row <- spec_summary |> filter(specimen_source_std == name)
  if (nrow(row)==0) return(c(label,"—","—","—"))
  c(label,
    as.character(row$n),
    cell2(row$n_pos, row$pct_pos),
    cell2(row$n_neg, row$pct_neg))
}

t2_rows <- list(
  c("Test Validity","","",""),
  c("  Total tests ordered",  as.character(n_total),"",""),
  c("  Analyzable (valid)",   as.character(n_valid),
    cell2(n_pos_all,pct_pos_all), cell2(n_neg_all,pct_neg_all)),
  c("  QNS",                  as.character(n_qns_n),"—",
    cell2(n_qns_n, pct_qns)),
  c("  Cancelled",            as.character(n_can),"—",
    cell2(n_can, pct_can)),
  c("Results by Assay (valid tests only)","","",""),
  ga("PIK3CA Multiplex"),
  ga("TEK"),
  ga("BRAF"),
  c("Specimen Source (valid tests only)","","",""),
  gs("Blood",           "  Blood"),
  gs("Lymphatic Fluid", "  Lymphatic Fluid"),
  gs("Other",           "  Other"),
  c("Specimen Collection Metrics","","",""),
  c("  Fluid volume, mL — median",
    sprintf("%.1f (N=%d)", med_vol,  vol_n),"",""),
  c("  DNA concentration, ng/µL — median",
    sprintf("%.1f (N=%d)", med_conc, conc_n),"",""),
  c("  DNA used in assay, ng — mean (SD)",
    sprintf("%s (N=%d)", fmt_ms(mean_used, sd_used), used_n),"",""),
  c("  EXON 21 WT droplets — mean (SD)",
    sprintf("%s (N=%d)", fmt_ms(mean_ex21, sd_ex21),
            sum(!is.na(df_valid$ex21_val))),"","")
)

t2_df <- bind_rows(lapply(t2_rows, function(r)
  tibble(Category=r[1], `Total n`=r[2],
         `Positive n (%)`=r[3], `Negative n (%)`=r[4])))

section2_rows <- which(t2_df$Category %in% c(
  "Test Validity","Results by Assay (valid tests only)",
  "Specimen Source (valid tests only)","Specimen Collection Metrics"))

t2_df |>
  kbl(caption = glue(
    "Table 2. Test Volume, Validity, and Results (N = {n_total} tests)"
  )) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) |>
  row_spec(section2_rows, bold = TRUE, background = "#D9E1F2")
```

*QNS = quantity not sufficient. Blank result entries excluded from all counts. NEG* treated as Negative.*

---

# Table 3 — Assay Performance Metrics

```{r table3-calcs}
#| label: table3-calcs

assay_list <- list(
  list(l="PIK3CA Multiplex", p="PIK3CA|PIK", ex=TRUE),
  list(l="TEK",              p="TEK",        ex=FALSE),
  list(l="BRAF",             p="BRAF",       ex=FALSE)
)

t3_rows <- list()

for (ai in assay_list) {
  sub <- df_valid |> filter(str_detect(assay_std, ai$p))
  nn  <- nrow(sub)

  # Specimen breakdown for this assay
  n_bl   <- sum(sub$specimen_source_std == "Blood",           na.rm=TRUE)
  n_ly   <- sum(sub$specimen_source_std == "Lymphatic Fluid", na.rm=TRUE)
  n_oth  <- sum(sub$specimen_source_std == "Other",           na.rm=TRUE)
  pct_bl  <- if(nn>0) n_bl/nn*100  else 0
  pct_ly  <- if(nn>0) n_ly/nn*100  else 0
  pct_oth <- if(nn>0) n_oth/nn*100 else 0

  for (res in c("Pos","Neg")) {
    s  <- if (res=="Pos") filter(sub, is_pos) else filter(sub, is_neg)
    sp <- list(
      blood = fmt_np(n_bl,  pct_bl),
      lymph = fmt_np(n_ly,  pct_ly),
      both  = fmt_np(n_oth, pct_oth)
    )

    n_res     <- nrow(s)
    pct_res   <- if(nn>0) n_res/nn*100 else 0

    t3_rows[[length(t3_rows)+1]] <- tibble(
      Assay                   = if (res=="Pos") ai$l else "",
      Result                  = res,
      `n (%)`                 = fmt_np(n_res, pct_res),
      `Fluid Vol (µL)`        = iqr_s(s$vol_val),
      `DNA Conc (ng/µL)`      = iqr_s(s$dna_conc_val),
      `DNA Used (ng)`         = msd_s(s$dna_used_val),
      `EXON 21 WT med(IQR)`   = if (ai$ex) iqr_s(s$ex21_val) else "N/A",
      `EXON 10 WT med(IQR)`   = if (ai$ex) iqr_s(s$ex10_val) else "N/A",
      `Blood Only n(%)`       = sp$blood,
      `Lymphatic Only n(%)`   = sp$lymph,
      `Other n(%)`            = sp$both
    )
  }
}

table3_df <- bind_rows(t3_rows)
```

```{r table3-display}
#| label: table3-display

section3_rows <- which(table3_df$Assay != "")

table3_df |>
  kbl(caption = "Table 3. Assay Performance Metrics by Result (valid tests only)") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) |>
  row_spec(section3_rows, bold = TRUE, background = "#D9E1F2") |>
  column_spec(1, bold = TRUE)
```

*IQR = interquartile range. SD = standard deviation. n (%) for Result column is proportion of that assay's valid tests. Specimen source columns reflect all valid tests for that assay regardless of result.*

---

# Table 4 — Clinical Impact of Test Results

```{r table4-calcs}
#| label: table4-calcs

# Restrict to valid tests with a result (POS or NEG)
df_t4 <- df_valid |> filter(is_pos | is_neg)

n_t4_pos <- sum(df_t4$is_pos)
n_t4_neg <- sum(df_t4$is_neg)

# Helper: count Y responses for a column, split by pos/neg
count_yn <- function(col, pos_filter, neg_filter) {
  pos_n <- sum(toupper(trimws(col[pos_filter])) == "Y", na.rm=TRUE)
  neg_n <- sum(toupper(trimws(col[neg_filter])) == "Y", na.rm=TRUE)
  pos_p <- if(n_t4_pos>0) pos_n/n_t4_pos*100 else 0
  neg_p <- if(n_t4_neg>0) neg_n/n_t4_neg*100 else 0
  c(fmt_np(pos_n, pos_p), fmt_np(neg_n, neg_p))
}

pos_idx <- df_t4$is_pos
neg_idx <- df_t4$is_neg

mgmt      <- count_yn(df_t4$mgmt_change,       pos_idx, neg_idx)
therapy_i <- count_yn(df_t4$therapy_initiated,  pos_idx, neg_idx)
therapy_m <- count_yn(df_t4$therapy_modified,   pos_idx, neg_idx)
procedure <- count_yn(df_t4$procedure_changed,  pos_idx, neg_idx)
add_test  <- count_yn(df_t4$additional_testing, pos_idx, neg_idx)

# No documented management change = mgmt_change == "N"
no_mgmt_pos <- sum(toupper(trimws(df_t4$mgmt_change[pos_idx])) == "N", na.rm=TRUE)
no_mgmt_neg <- sum(toupper(trimws(df_t4$mgmt_change[neg_idx])) == "N", na.rm=TRUE)
pct_no_pos  <- if(n_t4_pos>0) no_mgmt_pos/n_t4_pos*100 else 0
pct_no_neg  <- if(n_t4_neg>0) no_mgmt_neg/n_t4_neg*100 else 0

# Management documentation unavailable
unavail_pos <- sum(toupper(trimws(df_t4$mgmt_unavailable[pos_idx])) == "Y", na.rm=TRUE)
unavail_neg <- sum(toupper(trimws(df_t4$mgmt_unavailable[neg_idx])) == "Y", na.rm=TRUE)
pct_un_pos  <- if(n_t4_pos>0) unavail_pos/n_t4_pos*100 else 0
pct_un_neg  <- if(n_t4_neg>0) unavail_neg/n_t4_neg*100 else 0

t4 <- tibble(
  `Clinical Action / Outcome` = c(
    "Management Change Following Test Result",
    "  Any documented management change",
    "    Targeted therapy initiated (e.g., alpelisib, sirolimus)",
    "    Existing therapy modified or dose-adjusted",
    "    Surgical or procedural plan changed",
    "  No documented management change",
    "  Management documentation unavailable / not reported",
    "Patients who underwent additional testing, n (%)"
  ),
  `Positive Result (%)` = c(
    "", mgmt[1], therapy_i[1], therapy_m[1], procedure[1],
    fmt_np(no_mgmt_pos, pct_no_pos),
    fmt_np(unavail_pos, pct_un_pos),
    add_test[1]
  ),
  `Negative Result (%)` = c(
    "", mgmt[2], therapy_i[2], therapy_m[2], procedure[2],
    fmt_np(no_mgmt_neg, pct_no_neg),
    fmt_np(unavail_neg, pct_un_neg),
    add_test[2]
  )
)

section4_rows <- which(t4$`Clinical Action / Outcome` %in%
                         c("Management Change Following Test Result",
                           "Patients who underwent additional testing, n (%)"))
```

```{r table4-display}
#| label: table4-display

t4 |>
  kbl(caption = glue(
    "Table 4. Clinical Impact of Test Results ",
    "(Positive N = {n_t4_pos}; Negative N = {n_t4_neg})"
  )) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) |>
  row_spec(section4_rows, bold = TRUE, background = "#D9E1F2")
```

*Clinical actions documented by the ordering provider following return of test result. Includes all patients with a valid (analyzable) result. IQR = interquartile range. Management change determined by physician documentation in the medical record. Targeted therapies listed are examples; actual agents to be specified from chart review.*

---

# Data Accounting

```{r data-accounting-summary}
#| label: data-accounting-summary

tibble(
  Item  = c(
    "Total rows in raw file",
    "Spacer rows excluded (no Patient ID)",
    "Records retained for analysis",
    "Unique patients"
  ),
  Count = c(n_raw, n_spacer, n_records, n_pts)
) |>
  kbl(caption = "Data Accounting Summary") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

> **Missing data policy:** All records with a valid Patient ID are retained (N = `r n_records`). For each analysis, only rows with valid (non-QNS, non-cancelled, non-blank) results are included. Blank result entries (N = `r n_blanks`) are excluded from all counts. Age and sex N reflect records with non-missing values only.

```{r upload-to-drive}
#| label: upload-to-drive
#| include: false

# Upload rendered HTML to Google Drive outputs folder
output_file <- "spotseq_analysis_v2.html"
if (file.exists(output_file)) {
  drive_upload(
    media  = output_file,
    path   = as_id(drive_folder_id),
    name   = output_file,
    overwrite = TRUE
  )
}
```
