---
title: "SpotSeq Dataset Analysis"
author: "SpotSeq Analysis Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    theme: cosmo
params:
  export: false
execute:
  echo: false
  warning: false
  message: false
---

<!--
================================================================================
RENDERING MODES
================================================================================
Default (HTML preview only — fast, no exports):
    quarto preview spotseq_analysis_v2.qmd
    quarto render  spotseq_analysis_v2.qmd

Full export (HTML + Excel + 4 Word .docx files + Google Drive upload):
    quarto render  spotseq_analysis_v2.qmd -P export:true

Or in R:
    quarto::quarto_render("spotseq_analysis_v2.qmd",
                          execute_params = list(export = TRUE))

Outputs (when export=true) — saved to outputs/ in the project folder:
  - outputs/SpotSeq_Tables_<DATE>.xlsx           also uploaded to DRIVE_FOLDER_ID
  - outputs/Table[1-4]_<DATE>.docx
================================================================================
-->


```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(readxl)
library(janitor)
library(openxlsx)
library(gt)
library(glue)

data_path <- Sys.getenv("DATA_PATH")

stopifnot(
  "DATA_PATH not set in .Renviron" = nchar(data_path) > 0
)
```

```{r load-data}
#| label: load-data

# Auto-detect dataset file: DATA_FILE env var > most recent matching file
# This is robust to filename version bumps (e.g. _2.1 -> _2.2)
data_file <- Sys.getenv("DATA_FILE")
if (nchar(data_file) > 0) {
  data_file <- file.path(data_path, data_file)
} else {
  candidates <- list.files(
    data_path,
    pattern    = "SpotSeq.*deidentified.*\\.xlsx$",
    ignore.case = TRUE,
    full.names = TRUE
  )
  candidates <- candidates[!grepl("~\\$", candidates)]   # exclude Office lock files
  if (length(candidates) == 0)
    stop("No SpotSeq dataset file found in ", data_path,
         ". Set DATA_FILE in .Renviron or place the file in DATA_PATH.")
  data_file <- candidates[which.max(file.mtime(candidates))]
  message("Using dataset: ", basename(data_file))
}

# Read all columns as text first to capture displayed cell values, not formulas
raw <- read_excel(data_file, col_types = "text")

# Treat literal "." (a single period) as missing data.
# The source spreadsheet uses "." as an explicit placeholder for empty /
# inapplicable / "not documented" values in some cells. Convert to NA so
# downstream summaries don't count or mis-process these.
raw <- raw |>
  mutate(across(where(is.character),
                ~ if_else(trimws(.x) == ".", NA_character_, .x)))

df_raw    <- raw |> clean_names()
df        <- df_raw |> filter(!is.na(patient_id) & patient_id != "")
n_raw     <- nrow(raw)
n_spacer  <- n_raw - nrow(df)
n_records <- nrow(df)
n_pts     <- n_distinct(df$patient_id)
```

```{r specimen-standardise}
#| label: specimen-standardise

# Specimen source standardization: a hybrid pattern-based + exact-match
# classifier. Pattern-based rules handle the majority of variation
# (typos, anatomic site annotations, capitalization, word order); the
# exact-match list remains as a fallback for edge cases.
#
# Match priority (first match wins):
#   1. Blood-related patterns
#   2. Lymphatic-fluid-related patterns
#   3. Cyst-fluid patterns (non-lymphatic)
#   4. Other named anatomic sites (catch-all for non-lymphatic specimens)
#
# Anything that still doesn't match returns NA and is surfaced by the
# diagnostics-unmatched-spec-source chunk for review.

# Exact-match fallback lists — preserved for compatibility / explicit overrides
blood_vals <- toupper(c("Blood","Blood, venous","Lesional blood","Peripheral blood"))
lymph_vals <- toupper(c(
  "Lymphatic fluid","Lymphatic fluid, Site A","Lymphatic fluid, Site B",
  "Lymphatic fluid from left groin","Lymphatic cyst fluid",
  "Lymphatic cyst fluid from lymph node","Cyst fluid",
  "Lymphatic Malformation","Lymphatic malformation",
  "Chest, Lymphatic malformation"
))
other_vals <- toupper(c(
  "Aspirated Cyst Fluid","aspirated cyst fluid",
  "Abdomen, Lymphatic Cyst Fluid","Ankle, Left","Neck",
  "Tracheocutaneous fistula","Peritoneal fluid"
))

# Pattern-based classifier — uses regex on the uppercased string. The
# patterns are deliberately permissive: "LYMPATIC" catches the common
# typo "Lympatic fluid"; "PERIPHERAL.*BLOOD" catches "Peripheral blood";
# the anatomic-site terms catch non-lymphatic sites that should bin to "Other".
classify_specimen_source <- function(x_raw) {
  x <- toupper(trimws(x_raw))
  case_when(
    is.na(x) | x == ""                                    ~ NA_character_,
    # 1. Blood patterns: any "blood" descriptor (including typos via wildcard)
    grepl("BLOOD",  x)                                    ~ "Blood",
    # 2. Lymphatic patterns: "lymph" covers Lymphatic / Lympatic typo /
    #    lymphatic-malformation tissue / lymph node
    grepl("LYMPH",  x) | grepl("LYMPA", x)                ~ "Lymphatic Fluid",
    # 3. Cyst-fluid descriptors (non-lymphatic): "CYST" anywhere
    grepl("CYST",   x)                                    ~ "Other",
    # 4. Named non-lymphatic anatomic sites — extend list as new sites appear
    grepl(paste("ANKLE|NECK|EXTREMITY|FISTULA|PERITONE|",
                "ABDOMEN|CHEST|GROIN|FOOT|HAND|ARM|LEG|TRUNK",
                sep=""), x)                               ~ "Other",
    # 5. Exact-match fallback for anything not caught by patterns
    x %in% blood_vals                                     ~ "Blood",
    x %in% lymph_vals                                     ~ "Lymphatic Fluid",
    x %in% other_vals                                     ~ "Other",
    TRUE                                                  ~ NA_character_
  )
}

df <- df |>
  mutate(
    specimen_source_raw = toupper(trimws(specimen_source)),
    specimen_source_std = classify_specimen_source(specimen_source)
  )
```

```{r classify-results}
#| label: classify-results

df <- df |>
  mutate(
    # Standardise result: trim, upper
    result_clean   = toupper(trimws(result_pos_neg)),
    status_clean   = toupper(trimws(status)),

    # Classification flags
    is_qns        = status_clean %in% c("QNS","QNS FOR REPEAT","REPEAT") |
                    result_clean %in% c("QNS","QNS FOR REPEAT"),
    is_cancelled  = status_clean %in% c("CANCELLED","CANCELED") |
                    result_clean %in% c("CANCELLED","CANCELED"),

    # Blank result entries — omit from analysis
    is_blank      = is.na(result_pos_neg) | trimws(result_pos_neg) == "",

    # Valid = has a POS or NEG result, not QNS/cancelled/blank
    # status_clean == "COMPLETE" NOT required — some valid results have NA status
    is_valid      = !is_qns & !is_cancelled & !is_blank &
                    !is.na(result_clean) & result_clean != "" &
                    (grepl("^POS", result_clean) | grepl("^NEG", result_clean)),

    # Positive / Negative: NEG* and any capitalisation variant treated as NEG
    is_pos        = is_valid & grepl("^POS", result_clean),
    is_neg        = is_valid & grepl("^NEG", result_clean),   # catches NEG, NEG*, Neg, neg, etc.

    assay_std     = toupper(trimws(assay_ordered)),

    # Age: use displayed value from column 42 (age_at_date_of_collection_in_days)
    # Divide by 365.25 for decimal years — handles infants correctly
    age_days      = suppressWarnings(as.numeric(age_at_date_of_collection_in_days)),
    age_years     = age_days / 365.25,

    # Sex
    sex_val       = assigned_sex_at_birth,

    # Specimen type standardised (col 6)
    specimen_type_std = case_when(
      toupper(trimws(specimen_type)) == "BLOOD"     ~ "Blood",
      grepl("CYST", toupper(trimws(specimen_type))) ~ "Cyst Fluid",
      TRUE                                          ~ "Other"
    ),

    # Internal vs external patient (col 43)
    internal_val  = toupper(trimws(internal_patient_y_n)),

    # Volume / DNA / assay metrics — read from displayed values
    vol_val       = suppressWarnings(as.numeric(fluid_volume_extracted_converted)),
    dna_conc_val  = suppressWarnings(as.numeric(
                      dna_stock_ng_ul_blood_0_185ng_ul_reject_cyst_fluid_0_37ng_ul_reject)),
    dna_used_val  = suppressWarnings(as.numeric(total_ng_dna_used_in_assay)),
    ex21_val      = suppressWarnings(as.numeric(exon_21_wt_1500_avg_number_of_droplets)),
    ex10_val      = suppressWarnings(as.numeric(exon_10_wt_1500_avg_number_droplets)),

    # Table 4 — Clinical impact columns (44–49), exact clean_names() output
    mgmt_change        = toupper(trimws(any_documented_management_change_in_management_post_testing_y_n)),
    therapy_initiated  = toupper(trimws(targeted_therapy_initiated_post_testing_e_g_alpelisib_sirolimus_y_n)),
    therapy_modified   = toupper(trimws(existing_therapy_modified_or_dose_adjusted_post_testing_y_n_na)),
    procedure_changed  = toupper(trimws(surgical_or_procedural_plan_changed_post_testing_y_n_na)),
    additional_testing = toupper(trimws(patient_underwent_additional_testing_based_on_spot_seq_results_y_n_van)),
    mgmt_unavailable   = toupper(trimws(management_documentation_post_testing_unavailable_not_reported_y_n_na))
  )

# Assign each record to exactly one mutually exclusive category
# Priority: valid > qns > cancelled > unclassified
df <- df |>
  mutate(
    record_category = case_when(
      is_valid     ~ "valid",
      is_qns       ~ "qns",
      is_cancelled ~ "cancelled",
      is_blank     ~ "blank",
      TRUE         ~ "unclassified"
    )
  )

# Dynamically resolve column names that may vary between dataset versions
find_col <- function(df, pattern, label) {
  hits <- grep(pattern, names(df), value = TRUE, ignore.case = TRUE)
  if (length(hits) == 0) {
    message("WARNING: No column matching '", pattern, "' for ", label, " — will be NA.")
    return(NULL)
  }
  if (length(hits) > 1)
    message("NOTE: Multiple matches for '", pattern, "': ", paste(hits, collapse=", "),
            " — using: ", hits[1])
  hits[1]
}
col_single_probe <- find_col(df_raw, "single_probe",  "single_probe_val")
col_gene         <- find_col(df_raw, "^gene$",         "gene_val")
col_variant      <- find_col(df_raw, "^variant$",      "variant_val")
col_internal_vaf <- find_col(df_raw, "internal_vaf",   "internal_vaf_val")

# Per-cell VAF parser. Handles three input conventions seen in the source data:
#   1. Cells with explicit "%" sign (e.g., "5.2%", "130%") — already in percent units
#   2. Numeric > 1 without "%" (e.g., "5.2") — treat as already in percent units
#   3. Numeric ≤ 1 without "%" (e.g., "0.052", "0.131") — treat as decimal, ×100
# Per-cell parsing avoids the column-level scaling problem when the spreadsheet
# mixes conventions. Values >100% are passed through and surfaced by the
# diagnostic chunk for manual review (likely data entry errors).
parse_vaf_cell <- function(x) {
  if (is.na(x) || trimws(x) == "" || trimws(x) == ".") return(NA_real_)
  has_pct <- grepl("%", x)
  v <- suppressWarnings(as.numeric(gsub("%", "", x)))
  if (is.na(v)) return(NA_real_)
  if (has_pct) return(v)        # already in percent units
  if (v > 1)   return(v)        # >1 without % → already in percent units
  v * 100                       # ≤1 without % → decimal, scale to percent
}

df <- df |>
  mutate(
    single_probe_val = if (!is.null(col_single_probe))
                         suppressWarnings(as.numeric(.data[[col_single_probe]]))
                       else NA_real_,
    gene_val    = if (!is.null(col_gene))    toupper(trimws(.data[[col_gene]]))    else NA_character_,
    variant_val = if (!is.null(col_variant)) toupper(trimws(.data[[col_variant]])) else NA_character_,
    internal_vaf_val = if (!is.null(col_internal_vaf))
                         vapply(.data[[col_internal_vaf]], parse_vaf_cell, numeric(1))
                       else NA_real_
  )

# Filtered subsets — strictly mutually exclusive
df_valid    <- df |> filter(record_category == "valid")
df_qns      <- df |> filter(record_category == "qns")
df_cancel   <- df |> filter(record_category == "cancelled")
# Omit blanks from all counts
df_analysed <- df |> filter(record_category != "blank")
```

```{r data-reconciliation}
#| label: data-reconciliation

# Explicit accounting of all records — surfaces any unclassified entries
n_total    <- nrow(df_analysed)
n_valid    <- nrow(df_valid)
n_qns_n    <- nrow(df_qns)
n_can      <- nrow(df_cancel)
n_blanks   <- sum(df$record_category == "blank", na.rm = TRUE)
n_other    <- sum(df$record_category == "unclassified", na.rm = TRUE)
# All counts are strictly mutually exclusive via record_category
# n_valid + n_qns_n + n_can + n_other should equal n_total

tibble(
  Category = c(
    "Total rows in raw file",
    "Spacer rows excluded (no Patient ID)",
    "Records retained (with Patient ID)",
    "  Blank result entries (omitted from analysis)",
    "  Records included in analysis",
    "    Valid (analyzable)",
    "    QNS",
    "    Cancelled",
    "    Unclassified (investigate if > 0)"
  ),
  Count = c(
    n_raw, n_spacer, n_records,
    n_blanks, n_total,
    n_valid, n_qns_n, n_can, n_other
  )
) |>
  gt() |>
  tab_header(title = "Data Reconciliation — Record Accounting") |>
  tab_options(table.width = pct(50), column_labels.font.weight = "bold")
```

```{r patient-level}
#| label: patient-level

# Patient-level deduplication: ALL patients in the dataset (not just valid-test
# patients). This is the correct denominator for Table 1 cohort characteristics,
# which describe every patient submitted for testing — including those whose
# tests were QNS or cancelled.
#
# We use row_order (the order rows appear in the Excel file) as the dedup key
# because collection_date is stored as free-text in the source data and not
# reliably parseable. Row order corresponds to data entry order and is a
# well-defined, deterministic dedup key.

df_patients <- df |>
  mutate(row_order = row_number()) |>
  group_by(patient_id) |>
  slice_min(row_order, n = 1, with_ties = FALSE) |>
  ungroup()

n_pts_valid <- n_distinct(df_valid$patient_id)
```

```{r helpers}
#| label: helpers

fmt_ms  <- function(m, s) sprintf("%.1f (%.1f)", m, s)
fmt_med <- function(med, lo, hi) sprintf("%.1f [%.1f–%.1f]", med, lo, hi)
fmt_np  <- function(n, p) sprintf("%d (%.1f%%)", as.integer(n), p)
cell2   <- function(n, p) sprintf("%d (%.1f%%)", as.integer(n), p)
iqr_s   <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return("—")
  sprintf("%.1f [%.1f–%.1f]", median(x), quantile(x,.25), quantile(x,.75))
}
msd_s   <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return("—")
  sprintf("%.1f (%.1f)", mean(x), sd(x))
}
fmt_vaf_n <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  if (n == 0) return("—")
  sprintf("%.2f%% (n=%d)", mean(x), n)
}
fmt_p <- function(p) {
  if (is.na(p) || !is.finite(p)) return("--")
  if (p < 0.001) return("<0.001")
  sprintf("%.3f", p)
}
safe_wilcox <- function(x, y) {
  x <- x[!is.na(x)]; y <- y[!is.na(y)]
  if (length(x) < 2 || length(y) < 2) return(NA_real_)
  tryCatch(wilcox.test(x, y, exact = FALSE)$p.value, error = function(e) NA_real_)
}
safe_fisher <- function(grp, flag) {
  tab <- table(grp, flag)
  if (any(dim(tab) < 2)) return(NA_real_)
  tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
}
```

```{r table1-calcs}
#| label: table1-calcs

# Patient-level stats use df_patients (first test per unique sample_id_for_di_dataset)

# Age summary — using decimal years from AP column
age_n      <- sum(!is.na(df_patients$age_years))
age_mean   <- mean(df_patients$age_years,   na.rm = TRUE)
age_sd     <- sd(df_patients$age_years,     na.rm = TRUE)
age_med    <- median(df_patients$age_years, na.rm = TRUE)
age_q1     <- quantile(df_patients$age_years, 0.25, na.rm = TRUE)
age_q3     <- quantile(df_patients$age_years, 0.75, na.rm = TRUE)
age_min    <- min(df_patients$age_years,    na.rm = TRUE)
age_max    <- max(df_patients$age_years,    na.rm = TRUE)

# Pediatric / Adult split (< 18 years)
n_peds     <- sum(df_patients$age_years < 18,  na.rm = TRUE)
n_adult    <- sum(df_patients$age_years >= 18, na.rm = TRUE)
pct_peds   <- n_peds  / age_n * 100
pct_adult  <- n_adult / age_n * 100

# Sex
sex_clean  <- toupper(trimws(df_patients$sex_val))
n_female   <- sum(sex_clean == "F", na.rm = TRUE)
sex_n      <- sum(!is.na(df_patients$sex_val))
n_sex_n    <- sum(is.na(df_patients$sex_val))
pct_female <- n_female / sex_n * 100

# Internal vs External
internal_clean <- toupper(trimws(df_patients$internal_val))
n_internal     <- sum(internal_clean == "Y", na.rm = TRUE)
n_external     <- sum(internal_clean == "N", na.rm = TRUE)
n_int_ext      <- sum(!is.na(df_patients$internal_val) & df_patients$internal_val != "")
pct_int        <- if(n_int_ext > 0) n_internal / n_int_ext * 100 else 0
pct_ext        <- if(n_int_ext > 0) n_external / n_int_ext * 100 else 0

# Test-level internal count: how many of all tests (records) were performed
# on internally-processed patients. Distinct from the patient-level count above
# (n_internal = unique internal patients).
internal_clean_tests <- toupper(trimws(df$internal_val))
n_internal_tests     <- sum(internal_clean_tests == "Y", na.rm = TRUE)
pct_internal_tests   <- if (n_records > 0) n_internal_tests / n_records * 100 else 0
```

---

# Table 1 — Patient and Cohort Characteristics

```{r table1-display}
#| label: table1-display

t1 <- tibble(
  Characteristic = c(
    "Demographics",
    "  Age at collection, years",
    "    Mean (SD)",
    "    Median [IQR]",
    "    Range",
    "  Age group",
    "    Pediatric (< 18 years)",
    "    Adult (≥ 18 years)",
    "  Female sex assigned at birth, n (%)",
    "Patient Origin",
    "  Internally processed, n (%)",
    "  Externally referred, n (%)",
    "Testing Overview",
    "  Total tests ordered",
    "  Unique patients",
    "  Internally processed tests, n (%)"
  ),
  `N (%) / Value` = c(
    "",
    glue("(N = {age_n} of {n_pts} patients)"),
    fmt_ms(age_mean, age_sd),
    fmt_med(age_med, age_q1, age_q3),
    sprintf("%.2f – %.1f", age_min, age_max),
    "",
    fmt_np(n_peds,  pct_peds),
    fmt_np(n_adult, pct_adult),
    glue("{fmt_np(n_female, pct_female)} (N = {sex_n} of {n_pts} patients)"),
    glue("(N = {n_int_ext} of {n_pts} patients)"),
    fmt_np(n_internal, pct_int),
    fmt_np(n_external, pct_ext),
    "",
    as.character(n_records),
    as.character(n_pts),
    fmt_np(n_internal_tests, pct_internal_tests)
  )
)

section_rows <- which(t1$Characteristic %in%
                        c("Demographics","Patient Origin","Testing Overview"))

gt_t1 <- t1 |>
  gt() |>
  tab_header(
    title = glue("Table 1. Patient and Cohort Characteristics (N = {n_records} records; {n_pts} unique patients)")
  ) |>
  cols_label(Characteristic = "Characteristic", `N (%) / Value` = "n (%) / Value") |>
  tab_style(
    style = list(cell_fill(color="#D9E1F2"), cell_text(weight="bold")),
    locations = cells_body(rows = section_rows)
  ) |>
  tab_source_note(glue("SD = standard deviation. IQR = interquartile range. Age from column AP (days) ÷ 365.25. Patient-level rows (Demographics, Patient Origin) computed across all unique patients (N = {n_pts}); subgroup denominators exclude missing values. Internally processed tests = test-level count of all records ordered for internally managed patients.")) |>
  tab_options(table.width = pct(70), column_labels.font.weight = "bold")

gt_t1
```

---

# Table 2 — Test Volume, Validity, and Results

```{r table2-calcs}
#| label: table2-calcs

n_pos_all     <- sum(df_valid$is_pos, na.rm = TRUE)
n_neg_all     <- sum(df_valid$is_neg, na.rm = TRUE)
pct_pos_all   <- n_pos_all / n_valid * 100
pct_neg_all   <- n_neg_all / n_valid * 100
pct_qns       <- n_qns_n   / n_total * 100
pct_can       <- n_can     / n_total * 100

assay_summary <- df_valid |>
  mutate(assay_grp = case_when(
    str_detect(assay_std, "PIK3CA|PIK") ~ "PIK3CA Multiplex",
    str_detect(assay_std, "TEK")        ~ "TEK",
    str_detect(assay_std, "BRAF")       ~ "BRAF",
    TRUE                                ~ "Other"
  )) |>
  group_by(assay_grp) |>
  summarise(n=n(), n_pos=sum(is_pos), n_neg=sum(is_neg), .groups="drop") |>
  mutate(pct_pos = n_pos/n*100, pct_neg = n_neg/n*100)

spec_summary <- df_valid |>
  filter(!is.na(specimen_source_std)) |>
  group_by(specimen_source_std) |>
  summarise(n=n(), n_pos=sum(is_pos), n_neg=sum(is_neg), .groups="drop") |>
  mutate(pct_pos = n_pos/n*100, pct_neg = n_neg/n*100)

type_summary <- df_valid |>
  filter(!is.na(specimen_type_std)) |>
  group_by(specimen_type_std) |>
  summarise(n=n(), n_pos=sum(is_pos), n_neg=sum(is_neg), .groups="drop") |>
  mutate(pct_pos = n_pos/n*100, pct_neg = n_neg/n*100)

enc_both <- df_valid |>
  group_by(patient_id, collection_date) |>
  summarise(types = n_distinct(specimen_source_std), .groups="drop") |>
  filter(types > 1) |> nrow()

df_valid2 <- df_valid |>
  mutate(spec_grp = case_when(
    specimen_source_std == "Blood"           ~ "Blood only",
    specimen_source_std == "Lymphatic Fluid" ~ "Lymphatic fluid only",
    TRUE                                     ~ "Other"
  ))

n_blood_only <- sum(df_valid2$spec_grp == "Blood only",           na.rm = TRUE)
n_lymph_only <- sum(df_valid2$spec_grp == "Lymphatic fluid only", na.rm = TRUE)
n_both       <- enc_both
pct_blood    <- n_blood_only / n_valid * 100
pct_lymph    <- n_lymph_only / n_valid * 100
pct_both     <- n_both       / n_valid * 100

# Fluid volume uses df_patients (one entry per unique patient)
# Fluid volume: deduplicate df_valid by patient_id, exclude zeros and blanks
df_vol  <- df_valid |>
  filter(!is.na(vol_val) & vol_val > 0) |>
  group_by(patient_id) |>
  slice_min(row_number(), n = 1, with_ties = FALSE) |>
  ungroup()
vol_n   <- nrow(df_vol)
med_vol <- median(df_vol$vol_val, na.rm = TRUE)

# DNA and droplet metrics — all valid tests (test-level, no deduplication)
conc_n    <- sum(!is.na(df_valid$dna_conc_val))
used_n    <- sum(!is.na(df_valid$dna_used_val))
med_conc  <- median(df_valid$dna_conc_val, na.rm = TRUE)
# DNA used: report median [IQR] for consistency with Table 3
med_used  <- median(df_valid$dna_used_val,    na.rm = TRUE)
q1_used   <- quantile(df_valid$dna_used_val, 0.25, na.rm = TRUE)
q3_used   <- quantile(df_valid$dna_used_val, 0.75, na.rm = TRUE)
mean_ex21 <- mean(df_valid$ex21_val,       na.rm = TRUE)
sd_ex21   <- sd(df_valid$ex21_val,         na.rm = TRUE)
mean_ex10 <- mean(df_valid$ex10_val,       na.rm = TRUE)
sd_ex10   <- sd(df_valid$ex10_val,         na.rm = TRUE)

# Fluid volume pos/neg breakdown — unique sample_id_for_di_dataset, exclude zeros/blanks
df_vol_pos <- df_valid |> filter(is_pos, !is.na(vol_val), vol_val > 0) |>
              distinct(sample_id_for_di_dataset, .keep_all = TRUE)
df_vol_neg <- df_valid |> filter(is_neg, !is.na(vol_val), vol_val > 0) |>
              distinct(sample_id_for_di_dataset, .keep_all = TRUE)
med_vol_pos <- median(df_vol_pos$vol_val, na.rm = TRUE)
med_vol_neg <- median(df_vol_neg$vol_val, na.rm = TRUE)
vol_pos_n   <- nrow(df_vol_pos)
vol_neg_n   <- nrow(df_vol_neg)

# DNA/EXON pos/neg — all valid tests by result
med_conc_pos  <- median(df_valid$dna_conc_val[df_valid$is_pos], na.rm = TRUE)
med_conc_neg  <- median(df_valid$dna_conc_val[df_valid$is_neg], na.rm = TRUE)
# DNA used pos/neg: median [IQR]
med_used_pos  <- median(df_valid$dna_used_val[df_valid$is_pos],   na.rm = TRUE)
q1_used_pos   <- quantile(df_valid$dna_used_val[df_valid$is_pos], 0.25, na.rm = TRUE)
q3_used_pos   <- quantile(df_valid$dna_used_val[df_valid$is_pos], 0.75, na.rm = TRUE)
med_used_neg  <- median(df_valid$dna_used_val[df_valid$is_neg],   na.rm = TRUE)
q1_used_neg   <- quantile(df_valid$dna_used_val[df_valid$is_neg], 0.25, na.rm = TRUE)
q3_used_neg   <- quantile(df_valid$dna_used_val[df_valid$is_neg], 0.75, na.rm = TRUE)
mean_ex21_pos <- mean(df_valid$ex21_val[df_valid$is_pos],       na.rm = TRUE)
sd_ex21_pos   <- sd(df_valid$ex21_val[df_valid$is_pos],         na.rm = TRUE)
mean_ex21_neg <- mean(df_valid$ex21_val[df_valid$is_neg],       na.rm = TRUE)
sd_ex21_neg   <- sd(df_valid$ex21_val[df_valid$is_neg],         na.rm = TRUE)
mean_ex10_pos <- mean(df_valid$ex10_val[df_valid$is_pos],       na.rm = TRUE)
sd_ex10_pos   <- sd(df_valid$ex10_val[df_valid$is_pos],         na.rm = TRUE)
mean_ex10_neg <- mean(df_valid$ex10_val[df_valid$is_neg],       na.rm = TRUE)
sd_ex10_neg   <- sd(df_valid$ex10_val[df_valid$is_neg],         na.rm = TRUE)

ga <- function(name) {
  row <- assay_summary |> filter(assay_grp == name)
  if (nrow(row)==0) return(c(paste0("  ",name), "—","—","—"))
  c(paste0("  ",name),
    as.character(row$n),
    cell2(row$n_pos, row$pct_pos),
    cell2(row$n_neg, row$pct_neg))
}

gtype <- function(name, label) {
  row <- type_summary |> filter(specimen_type_std == name)
  if (nrow(row)==0) return(c(label,"—","—","—"))
  c(label,
    as.character(row$n),
    cell2(row$n_pos, row$pct_pos),
    cell2(row$n_neg, row$pct_neg))
}

gs <- function(name, label) {
  row <- spec_summary |> filter(specimen_source_std == name)
  if (nrow(row)==0) return(c(label,"—","—","—"))
  c(label,
    as.character(row$n),
    cell2(row$n_pos, row$pct_pos),
    cell2(row$n_neg, row$pct_neg))
}

t2_rows <- list(
  c("Test Validity","","",""),
  c("  Total tests ordered",  as.character(n_total),"",""),
  c("  Analyzable (valid)", fmt_np(n_valid, n_valid/n_records*100),
    cell2(n_pos_all,pct_pos_all), cell2(n_neg_all,pct_neg_all)),
  c("  QNS", fmt_np(n_qns_n, n_qns_n/n_records*100), "", ""),
  c("  Cancelled", fmt_np(n_can, n_can/n_records*100), "", ""),
  c("Results by Assay (valid tests only)","","",""),
  ga("PIK3CA Multiplex"),
  ga("TEK"),
  ga("BRAF"),
  c("Specimen Type (valid tests only)","","",""),
  gtype("Blood",        "  Blood"),
  gtype("Cyst Fluid",   "  Cyst Fluid"),
  gtype("Other", "  FFPE Scrolls"),
  c("Specimen Source (valid tests only)","","",""),
  gs("Blood",           "  Blood"),
  gs("Lymphatic Fluid", "  Lymphatic Fluid"),
  gs("Other",           "  Other"),
  c("Specimen Collection Metrics","","",""),
  c(sprintf("  Fluid volume, mL — median (N=%d patients)", vol_n),
    sprintf("%.1f", med_vol),
    sprintf("%.1f (N=%d)", med_vol_pos, vol_pos_n),
    sprintf("%.1f (N=%d)", med_vol_neg, vol_neg_n)),
  c(sprintf("  DNA concentration, ng/µL — median (N=%d)", conc_n),
    sprintf("%.1f", med_conc),
    sprintf("%.1f", med_conc_pos),
    sprintf("%.1f", med_conc_neg)),
  c(sprintf("  DNA used in assay, ng — median [IQR] (N=%d)", used_n),
    fmt_med(med_used, q1_used, q3_used),
    fmt_med(med_used_pos, q1_used_pos, q3_used_pos),
    fmt_med(med_used_neg, q1_used_neg, q3_used_neg)),
  c(sprintf("  EXON 21 WT droplets — mean (SD) (N=%d)", sum(!is.na(df_valid$ex21_val))),
    fmt_ms(mean_ex21, sd_ex21),
    fmt_ms(mean_ex21_pos, sd_ex21_pos),
    fmt_ms(mean_ex21_neg, sd_ex21_neg)),
  c(sprintf("  EXON 10 WT droplets — mean (SD) (N=%d)", sum(!is.na(df_valid$ex10_val))),
    fmt_ms(mean_ex10, sd_ex10),
    fmt_ms(mean_ex10_pos, sd_ex10_pos),
    fmt_ms(mean_ex10_neg, sd_ex10_neg))
)

t2_df <- bind_rows(lapply(t2_rows, function(r)
  tibble(Category=r[1], `Total n`=r[2],
         `Positive n (%)`=r[3], `Negative n (%)`=r[4])))

section2_rows <- which(t2_df$Category %in% c(
  "Test Validity","Results by Assay (valid tests only)",
  "Specimen Type (valid tests only)",
  "Specimen Source (valid tests only)","Specimen Collection Metrics"))

gt_t2 <- t2_df |>
  gt() |>
  tab_header(
    title    = "Table 2. Test Volume, Validity, and Results",
    subtitle = glue("N = {n_total} tests")
  ) |>
  cols_label(
    Category       = "Category",
    `Total n`      = "n / Value",
    `Positive n (%)` = "Positive n (%)",
    `Negative n (%)` = "Negative n (%)"
  ) |>
  tab_style(
    style = list(cell_fill(color="#D9E1F2"), cell_text(weight="bold")),
    locations = cells_body(rows = section2_rows)
  ) |>
  tab_source_note("QNS = quantity not sufficient. Blank result entries excluded from all counts. NEG* treated as Negative. Fluid volume reported at patient level (unique patient ID); DNA and droplet metrics at test level (all valid tests). Positive/negative fluid volume uses unique sample IDs.") |>
  tab_options(table.width = pct(80), column_labels.font.weight = "bold")

gt_t2
```

---

# Table 3 — Assay Performance Metrics

```{r table3-calcs}
#| label: table3-calcs

# PIK3CA variant sub-rows (GNAQ excluded per protocol; TEK/BRAF renamed inline)
# Patterns include protein (p.E542K), HGVS protein-3-letter (p.GLU542LYS), and
# HGVS coding (c.1624G>A) — robust to whichever notation is entered in the spreadsheet.
pik3ca_variants <- list(
  list(label = "  p.E542K",  patterns = c("E542K",  "GLU542LYS", "1624G>A")),
  list(label = "  p.E545K",  patterns = c("E545K",  "GLU545LYS", "1633G>A")),
  list(label = "  p.H1047R", patterns = c("H1047R", "HIS1047ARG","3140A>G"))
)
# Helper: TRUE for rows whose variant_val (or gene_val) matches any pattern
match_variant <- function(df, patterns) {
  hits <- rep(FALSE, nrow(df))
  for (p in patterns) {
    hits <- hits |
      grepl(p, df$variant_val, ignore.case = TRUE) |
      grepl(p, df$gene_val,    ignore.case = TRUE)
  }
  hits
}

assay_list <- list(
  list(l="PIK3CA Multiplex", p="PIK3CA|PIK",
       pos_label="Positive",           neg_label="Negative"),
  list(l="TEK",              p="TEK",
       pos_label="Positive (p.L914F)", neg_label="Negative"),
  list(l="BRAF",             p="BRAF",
       pos_label="Positive (p.V600E)", neg_label="Negative")
)

# Specimen type counts computed per result group (not whole-assay totals)
spec_counts_r <- function(s) {
  n_r  <- nrow(s)
  n_bl <- sum(s$specimen_type_std == "Blood",      na.rm = TRUE)
  n_ly <- sum(s$specimen_type_std == "Cyst Fluid", na.rm = TRUE)
  list(
    bl = fmt_np(n_bl, if (n_r > 0) n_bl / n_r * 100 else 0),
    ly = fmt_np(n_ly, if (n_r > 0) n_ly / n_r * 100 else 0)
  )
}

t3_rows <- list()

for (ai in assay_list) {
  sub   <- df_valid |> filter(str_detect(assay_std, ai$p))
  nn    <- nrow(sub)
  s_pos <- filter(sub, is_pos)
  s_neg <- filter(sub, is_neg)

  # --- Assay header row ---
  t3_rows[[length(t3_rows)+1]] <- tibble(
    Assay                              = ai$l,
    Result                             = "",
    `n (%)`                            = "",
    `Fluid Vol (mL) Median [IQR]`      = "",
    `DNA Conc (ng/µL) Median [IQR]`    = "",
    `DNA Used (ng) Median [IQR]`       = "",
    `Average Internal VAF % (n)`       = "",
    `Specimen Type: Blood n (%)`       = "",
    `Specimen Type: Cyst Fluid n (%)`  = ""
  )

  # --- Positive row ---
  n_pos   <- nrow(s_pos)
  pct_pos <- if (nn > 0) n_pos / nn * 100 else 0
  s_vol_p <- s_pos |> filter(!is.na(vol_val) & vol_val > 0) |>
             distinct(sample_id_for_di_dataset, .keep_all = TRUE)
  sp_p    <- spec_counts_r(s_pos)

  t3_rows[[length(t3_rows)+1]] <- tibble(
    Assay                              = "",
    Result                             = ai$pos_label,
    `n (%)`                            = fmt_np(n_pos, pct_pos),
    `Fluid Vol (mL) Median [IQR]`      = iqr_s(s_vol_p$vol_val),
    `DNA Conc (ng/µL) Median [IQR]`    = iqr_s(s_pos$dna_conc_val),
    `DNA Used (ng) Median [IQR]`       = iqr_s(s_pos$dna_used_val),
    `Average Internal VAF % (n)`       = fmt_vaf_n(s_pos$internal_vaf_val),
    `Specimen Type: Blood n (%)`       = sp_p$bl,
    `Specimen Type: Cyst Fluid n (%)`  = sp_p$ly
  )

  # --- PIK3CA variant sub-rows ---
  if (ai$l == "PIK3CA Multiplex") {
    for (v in pik3ca_variants) {
      sv      <- s_pos |> filter(match_variant(s_pos, v$patterns))
      n_sv    <- nrow(sv)
      pct_sv  <- if (n_pos > 0) n_sv / n_pos * 100 else 0
      s_vol_v <- sv |> filter(!is.na(vol_val) & vol_val > 0) |>
                 distinct(sample_id_for_di_dataset, .keep_all = TRUE)
      sp_v    <- spec_counts_r(sv)
      t3_rows[[length(t3_rows)+1]] <- tibble(
        Assay                              = "",
        Result                             = v$label,
        `n (%)`                            = fmt_np(n_sv, pct_sv),
        `Fluid Vol (mL) Median [IQR]`      = iqr_s(s_vol_v$vol_val),
        `DNA Conc (ng/µL) Median [IQR]`    = iqr_s(sv$dna_conc_val),
        `DNA Used (ng) Median [IQR]`       = iqr_s(sv$dna_used_val),
        `Average Internal VAF % (n)`       = fmt_vaf_n(sv$internal_vaf_val),
        `Specimen Type: Blood n (%)`       = sp_v$bl,
        `Specimen Type: Cyst Fluid n (%)`  = sp_v$ly
      )
    }
  }

  # --- Negative row ---
  n_neg   <- nrow(s_neg)
  pct_neg <- if (nn > 0) n_neg / nn * 100 else 0
  s_vol_n <- s_neg |> filter(!is.na(vol_val) & vol_val > 0) |>
             distinct(sample_id_for_di_dataset, .keep_all = TRUE)
  sp_n    <- spec_counts_r(s_neg)

  t3_rows[[length(t3_rows)+1]] <- tibble(
    Assay                              = "",
    Result                             = ai$neg_label,
    `n (%)`                            = fmt_np(n_neg, pct_neg),
    `Fluid Vol (mL) Median [IQR]`      = iqr_s(s_vol_n$vol_val),
    `DNA Conc (ng/µL) Median [IQR]`    = iqr_s(s_neg$dna_conc_val),
    `DNA Used (ng) Median [IQR]`       = iqr_s(s_neg$dna_used_val),
    `Average Internal VAF % (n)`       = fmt_vaf_n(s_neg$internal_vaf_val),
    `Specimen Type: Blood n (%)`       = sp_n$bl,
    `Specimen Type: Cyst Fluid n (%)`  = sp_n$ly
  )

  # --- p-value row (Mann-Whitney U / Fisher's exact, Positive vs Negative) ---
  # VAF is not compared (negatives by definition have no detected variant).
  p_vol  <- fmt_p(safe_wilcox(s_vol_p$vol_val, s_vol_n$vol_val))
  p_conc <- fmt_p(safe_wilcox(s_pos$dna_conc_val,  s_neg$dna_conc_val))
  p_used <- fmt_p(safe_wilcox(s_pos$dna_used_val,  s_neg$dna_used_val))
  sp_sub <- sub |> filter(specimen_type_std %in% c("Blood", "Cyst Fluid"))
  p_spec <- if (nrow(sp_sub) >= 2)
              fmt_p(safe_fisher(sp_sub$specimen_type_std, sp_sub$is_pos))
            else "--"

  t3_rows[[length(t3_rows)+1]] <- tibble(
    Assay                              = "",
    Result                             = "p-value",
    `n (%)`                            = "",
    `Fluid Vol (mL) Median [IQR]`      = p_vol,
    `DNA Conc (ng/µL) Median [IQR]`    = p_conc,
    `DNA Used (ng) Median [IQR]`       = p_used,
    `Average Internal VAF % (n)`       = "—",
    `Specimen Type: Blood n (%)`       = p_spec,
    `Specimen Type: Cyst Fluid n (%)`  = ""
  )
}

table3_df    <- bind_rows(t3_rows)
```

```{r table3-display}
#| label: table3-display

section3_rows <- which(table3_df$Assay != "")
pval3_rows    <- which(table3_df$Result == "p-value")
var3_rows     <- which(grepl("^  p\\.", table3_df$Result))

gt_t3 <- table3_df |>
  gt() |>
  tab_header(
    title    = "Table 3. Assay Performance Metrics by Result",
    subtitle = "Valid tests only"
  ) |>
  tab_style(
    style = list(cell_fill(color = "#D9E1F2"), cell_text(weight = "bold")),
    locations = cells_body(rows = section3_rows)
  ) |>
  tab_style(
    style = list(cell_text(style = "italic", color = "#595959")),
    locations = cells_body(rows = pval3_rows)
  ) |>
  tab_style(
    style = cell_text(indent = px(16)),
    locations = cells_body(columns = Result, rows = var3_rows)
  ) |>
  tab_source_note(paste(
    "IQR = interquartile range. VAF = variant allele frequency.",
    "Continuous metrics reported as Median [IQR]; VAF reported as mean (n contributing).",
    "Specimen type breakdown reflects tests within each result group.",
    "p-values: Mann-Whitney U test for continuous variables;",
    "Fisher's exact test for specimen type (Blood vs Cyst Fluid).",
    "VAF not statistically compared (negatives have no detected variant by definition).",
    "PIK3CA variant sub-row percentages are of PIK3CA Positive tests."
  )) |>
  tab_options(table.width = pct(100), column_labels.font.weight = "bold")

gt_t3
```

---

# Table 4 — Clinical Impact of Test Results

```{r table4-calcs}
#| label: table4-calcs

# Restrict to valid tests with POS or NEG result
df_t4 <- df_valid |> filter(is_pos | is_neg)

# N for Table 4 = total valid tests (pos or neg) with non-blank mgmt_change
# n (%) in each row = count of Y / N_total
all_mgmt   <- toupper(trimws(df_t4$mgmt_change))
n_t4_total <- sum(!is.na(all_mgmt) & all_mgmt != "" & all_mgmt != "NA")

# Helper: count Y (or specified values) across ALL valid tests, % out of n_t4_total
count_yn <- function(col, values_to_count = "Y") {
  vals      <- toupper(trimws(col))
  non_blank <- vals[!is.na(vals) & vals != "" & vals != "NA"]
  n_yes     <- sum(non_blank %in% values_to_count)
  pct       <- if (n_t4_total > 0) n_yes / n_t4_total * 100 else 0
  list(n = n_yes, fmt = fmt_np(n_yes, pct))
}

# Any management change — Y only
mgmt_res <- count_yn(df_t4$mgmt_change)

# Targeted therapy initiated — Y only
ti_res   <- count_yn(df_t4$therapy_initiated)

# Therapy modified — Y only
tm_res   <- count_yn(df_t4$therapy_modified)

# Procedure changed — Y only
pc_res   <- count_yn(df_t4$procedure_changed)

# No management change — N only
no_res   <- count_yn(df_t4$mgmt_change, values_to_count = "N")

# Mgmt unavailable — Y only
un_res   <- count_yn(df_t4$mgmt_unavailable)

# Additional testing — Y or VAN
at_res   <- count_yn(df_t4$additional_testing, values_to_count = c("Y","VAN"))

t4 <- tibble(
  `Clinical Action / Outcome` = c(
    "  Any documented management change",
    "    Targeted therapy initiated (e.g., alpelisib, sirolimus)",
    "    Existing therapy modified or dose-adjusted",
    "    Surgical or procedural plan changed",
    "  No documented management change",
    "  Management documentation unavailable / not reported",
    "Patients who underwent additional testing"
  ),
  `n (%)` = c(
    mgmt_res$fmt, ti_res$fmt, tm_res$fmt, pc_res$fmt,
    no_res$fmt, un_res$fmt, at_res$fmt
  )
)

section4_rows <- integer(0)  # no highlighted rows in Table 4
```

```{r table4-display}
#| label: table4-display

gt_t4 <- t4 |>
  gt() |>
  tab_header(title = glue("Table 4. Clinical Impact of Test Results (Total patients assessed, N = {n_t4_total})")) |>
  tab_style(
    style = list(cell_fill(color="#D9E1F2"), cell_text(weight="bold")),
    locations = cells_body(rows = section4_rows)
  ) |>
  tab_source_note("Clinical actions were assessed in internally managed patients with valid (non-QNS, non-cancelled) results, as complete clinical documentation was available only for this subgroup. Y and VAN counted as positive for additional testing. Blank entries excluded from denominators. Targeted therapies listed are examples; actual agents confirmed by chart review.") |>
  tab_options(table.width = pct(70), column_labels.font.weight = "bold")

gt_t4
```

---

# Pre-analytical Correlation Analyses

Supplementary analyses examining whether specimen-level pre-analytical factors
(fluid volume, QNS classification) are associated with assay output. Restricted
to PIK3CA Multiplex valid tests.

```{r preanalytical-spearman}
#| label: preanalytical-spearman

# Spearman rank correlation between fluid volume and wild-type droplet counts,
# stratified by specimen type, within PIK3CA Multiplex valid tests only.
#
# Rationale for choices:
#   - Spearman (not Pearson): both variables are non-normally distributed and
#     span multiple orders of magnitude; Spearman captures monotonic
#     relationships of any shape without distributional assumptions.
#   - Restricted to PIK3CA: TEK and BRAF have too few valid tests for a
#     meaningful correlation analysis.
#   - Stratified by specimen type: blood and cyst fluid have different
#     pre-analytical pathways; combining them would introduce confounding.

# Build the analytic subset
df_corr <- df_valid |>
  filter(str_detect(assay_std, "PIK3CA|PIK")) |>
  filter(!is.na(vol_val), vol_val > 0)

spearman_row <- function(label, x, y) {
  x <- x[!is.na(x) & !is.na(y)]
  y <- y[!is.na(x) & !is.na(y)]
  if (length(x) < 4) {
    return(tibble(Comparison = label, n = length(x),
                  `Spearman ρ` = "—", `p-value` = "—"))
  }
  ct <- suppressWarnings(
    cor.test(x, y, method = "spearman", exact = FALSE))
  tibble(Comparison = label,
         n           = length(x),
         `Spearman ρ` = sprintf("%.3f", ct$estimate),
         `p-value`    = fmt_p(ct$p.value))
}

corr_results <- bind_rows(
  # Cyst Fluid only
  spearman_row("Cyst Fluid: Volume vs EXON 21 WT droplets",
               df_corr$vol_val[df_corr$specimen_type_std == "Cyst Fluid"],
               df_corr$ex21_val[df_corr$specimen_type_std == "Cyst Fluid"]),
  spearman_row("Cyst Fluid: Volume vs EXON 10 WT droplets",
               df_corr$vol_val[df_corr$specimen_type_std == "Cyst Fluid"],
               df_corr$ex10_val[df_corr$specimen_type_std == "Cyst Fluid"]),
  # Blood only
  spearman_row("Blood: Volume vs EXON 21 WT droplets",
               df_corr$vol_val[df_corr$specimen_type_std == "Blood"],
               df_corr$ex21_val[df_corr$specimen_type_std == "Blood"]),
  spearman_row("Blood: Volume vs EXON 10 WT droplets",
               df_corr$vol_val[df_corr$specimen_type_std == "Blood"],
               df_corr$ex10_val[df_corr$specimen_type_std == "Blood"])
)

gt_corr <- corr_results |>
  gt() |>
  tab_header(
    title    = "Spearman correlations: fluid volume vs. wild-type droplet counts",
    subtitle = "PIK3CA Multiplex valid tests, stratified by specimen type"
  ) |>
  tab_source_note(paste(
    "Spearman rank correlation (non-parametric, distribution-free).",
    "Restricted to specimens with fluid volume > 0.",
    "Correlations not reported where n < 4."
  )) |>
  tab_options(table.width = pct(85), column_labels.font.weight = "bold")

gt_corr
```

```{r preanalytical-scatter}
#| label: preanalytical-scatter
#| fig-width: 9
#| fig-height: 7

# Scatter plot: fluid volume vs wild-type droplets, faceted by specimen type
# and droplet target, with LOESS smoother. Log-log axes because both variables
# span several orders of magnitude in real data.
#
# Reshape to long form to facet both EXON targets in one plot.
df_scatter <- df_corr |>
  select(specimen_type_std, vol_val, ex21_val, ex10_val) |>
  filter(specimen_type_std %in% c("Blood", "Cyst Fluid")) |>
  tidyr::pivot_longer(
    cols      = c(ex21_val, ex10_val),
    names_to  = "target",
    values_to = "wt_droplets"
  ) |>
  mutate(target = recode(target,
                         ex21_val = "EXON 21 WT droplets",
                         ex10_val = "EXON 10 WT droplets")) |>
  filter(!is.na(wt_droplets), wt_droplets > 0)

p_spearman_scatter <- ggplot(df_scatter, aes(x = vol_val, y = wt_droplets)) +
  geom_point(alpha = 0.5, size = 1.8, colour = "#2C5282") +
  geom_smooth(method = "loess", se = TRUE,
              colour = "#C53030", fill = "#FED7D7", linewidth = 0.8) +
  scale_x_log10() +
  scale_y_log10() +
  facet_grid(target ~ specimen_type_std, scales = "free_y") +
  labs(
    x = "Fluid volume (mL, log scale)",
    y = "Wild-type droplets (log scale)",
    title    = "Fluid volume vs. wild-type droplet counts",
    subtitle = "PIK3CA Multiplex valid tests, by specimen type (LOESS smoother with 95% CI)"
  ) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold"))

p_spearman_scatter
```

```{r preanalytical-qns-vs-valid}
#| label: preanalytical-qns-vs-valid

# Compare fluid volume between QNS and valid specimens.
# Clinically meaningful question: are low-volume specimens at higher risk of
# failing the assay (returning QNS rather than a valid result)?
# Mann-Whitney U test (non-parametric, two-sided).
df_qv <- df |>
  filter(record_category %in% c("valid", "qns"),
         !is.na(vol_val), vol_val > 0) |>
  mutate(record_category = factor(record_category,
                                  levels = c("valid", "qns"),
                                  labels = c("Valid", "QNS")))

qv_summary <- df_qv |>
  group_by(`Result Category` = record_category) |>
  summarize(
    n            = n(),
    `Median Volume (mL)` = sprintf("%.1f", median(vol_val)),
    `IQR`        = sprintf("[%.1f–%.1f]",
                           quantile(vol_val, 0.25),
                           quantile(vol_val, 0.75)),
    `Range`      = sprintf("%.1f – %.1f", min(vol_val), max(vol_val)),
    .groups = "drop"
  )

qv_p <- if (n_distinct(df_qv$record_category) == 2 && nrow(df_qv) >= 4) {
  fmt_p(suppressWarnings(
    wilcox.test(vol_val ~ record_category, data = df_qv, exact = FALSE)$p.value))
} else {
  "—"
}

gt_qv <- qv_summary |>
  gt() |>
  tab_header(
    title    = "Fluid volume: QNS vs. valid specimens",
    subtitle = glue("Mann–Whitney U test: p = {qv_p}")
  ) |>
  tab_source_note(paste(
    "Restricted to specimens with recorded fluid volume > 0.",
    "Two-sided Mann–Whitney U test (Wilcoxon rank-sum).",
    "A lower median volume in the QNS group would suggest that low-volume",
    "specimens are at higher risk of failing the assay."
  )) |>
  tab_options(table.width = pct(85), column_labels.font.weight = "bold")

gt_qv
```

```{r preanalytical-qns-boxplot}
#| label: preanalytical-qns-boxplot
#| fig-width: 6
#| fig-height: 4.5

p_qns_pooled <- ggplot(df_qv, aes(x = record_category, y = vol_val, fill = record_category)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.75) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.5, size = 1.8) +
  scale_y_log10() +
  scale_fill_manual(values = c("Valid" = "#2C5282", "QNS" = "#C53030")) +
  labs(
    x = NULL,
    y = "Fluid volume (mL, log scale)",
    title    = "Fluid volume by result category",
    subtitle = "QNS vs. valid specimens"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))

p_qns_pooled
```

```{r preanalytical-qns-rate-by-quartile}
#| label: preanalytical-qns-rate-by-quartile

# QNS rate by fluid-volume quartile. Reframes the QNS-vs-valid comparison
# as a clinically actionable threshold question: at what volume range does
# QNS risk become substantial? Quartiles are computed across the COMBINED
# valid + QNS pool (so each bin contains roughly equal n), then QNS rate is
# computed within each bin.
df_q <- df |>
  filter(record_category %in% c("valid", "qns"),
         !is.na(vol_val), vol_val > 0) |>
  mutate(vol_q = ntile(vol_val, 4))

# Per-quartile summary
quartile_summary <- df_q |>
  group_by(vol_q) |>
  summarize(
    n_total       = n(),
    n_qns         = sum(record_category == "qns"),
    qns_rate      = n_qns / n_total * 100,
    vol_min       = min(vol_val),
    vol_max       = max(vol_val),
    vol_median    = median(vol_val),
    .groups       = "drop"
  ) |>
  mutate(
    `Volume Quartile` = paste0("Q", vol_q, " (", sprintf("%.1f–%.1f mL", vol_min, vol_max), ")"),
    `n (Valid + QNS)` = as.character(n_total),
    `QNS, n (%)`      = sprintf("%d (%.1f%%)", n_qns, qns_rate),
    `Median Volume (mL)` = sprintf("%.1f", vol_median)
  ) |>
  select(`Volume Quartile`, `n (Valid + QNS)`, `QNS, n (%)`, `Median Volume (mL)`)

# Test for linear trend in QNS rate across quartiles
trend_p <- tryCatch({
  fmt_p(prop.trend.test(
    df_q |> group_by(vol_q) |> summarize(x = sum(record_category == "qns"),
                                          .groups = "drop") |> pull(x),
    df_q |> group_by(vol_q) |> summarize(n = n(),
                                          .groups = "drop") |> pull(n)
  )$p.value)
}, error = function(e) "—")

gt_quartile <- quartile_summary |>
  gt() |>
  tab_header(
    title    = "QNS rate by fluid-volume quartile",
    subtitle = glue("Cochran–Armitage trend test: p = {trend_p}")
  ) |>
  tab_source_note(paste(
    "Quartiles defined on the combined valid + QNS pool (n ≈ equal per bin).",
    "Trend test assesses whether QNS proportion decreases monotonically with",
    "increasing volume. Useful for setting a clinical minimum-volume recommendation."
  )) |>
  tab_options(table.width = pct(85), column_labels.font.weight = "bold")

gt_quartile
```

```{r preanalytical-qns-by-spectype}
#| label: preanalytical-qns-by-spectype

# QNS-vs-valid volume comparison, stratified by specimen type.
# More directly clinically actionable than the pooled analysis because
# specimen types differ in expected volume and have type-specific
# collection workflows.

qns_by_spec <- df |>
  filter(record_category %in% c("valid", "qns"),
         !is.na(vol_val), vol_val > 0,
         specimen_type_std %in% c("Blood", "Cyst Fluid")) |>
  mutate(record_category = factor(record_category,
                                  levels = c("valid", "qns"),
                                  labels = c("Valid", "QNS")))

# Summary stats by specimen type and result category
qns_spec_summary <- qns_by_spec |>
  group_by(`Specimen Type` = specimen_type_std,
           `Result Category` = record_category) |>
  summarize(
    n               = n(),
    `Median Volume (mL)` = sprintf("%.1f", median(vol_val)),
    `IQR`           = sprintf("[%.1f–%.1f]",
                              quantile(vol_val, 0.25),
                              quantile(vol_val, 0.75)),
    .groups = "drop"
  )

# Mann-Whitney U p-value within each specimen type
spec_pvalues <- qns_by_spec |>
  group_by(specimen_type_std) |>
  summarize(
    n_valid = sum(record_category == "Valid"),
    n_qns   = sum(record_category == "QNS"),
    p       = if (n_valid >= 2 && n_qns >= 2) {
                fmt_p(suppressWarnings(wilcox.test(vol_val ~ record_category,
                                                   exact = FALSE)$p.value))
              } else "— (insufficient n)",
    .groups = "drop"
  ) |>
  mutate(label = glue("{specimen_type_std}: p = {p} ",
                      "(valid n={n_valid}, QNS n={n_qns})")) |>
  pull(label)

gt_qns_spec <- qns_spec_summary |>
  gt(groupname_col = "Specimen Type") |>
  tab_header(
    title    = "Fluid volume: QNS vs. valid, by specimen type",
    subtitle = paste(spec_pvalues, collapse = " | ")
  ) |>
  tab_source_note(paste(
    "Two-sided Mann–Whitney U test within each specimen type.",
    "Restricted to specimens with fluid volume > 0.",
    "Stratified analysis avoids confounding by specimen-type-dependent",
    "differences in expected volume and collection workflow."
  )) |>
  tab_options(table.width = pct(85), column_labels.font.weight = "bold")

gt_qns_spec
```

```{r preanalytical-qns-by-spectype-plot}
#| label: preanalytical-qns-by-spectype-plot
#| fig-width: 8
#| fig-height: 4.5

p_qns_spectype <- ggplot(qns_by_spec, aes(x = record_category, y = vol_val,
                         fill = record_category)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.75) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.5, size = 1.8) +
  facet_wrap(~ specimen_type_std, nrow = 1) +
  scale_y_log10() +
  scale_fill_manual(values = c("Valid" = "#2C5282", "QNS" = "#C53030")) +
  labs(
    x = NULL,
    y = "Fluid volume (mL, log scale)",
    title    = "Fluid volume by result category, stratified by specimen type",
    subtitle = "QNS vs. valid specimens within Blood and Cyst Fluid subgroups"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold"))

p_qns_spectype
```

```{r preanalytical-qns-logreg}
#| label: preanalytical-qns-logreg

# Logistic regression: P(QNS) ~ log(volume) + specimen_type
#
# Models the probability of a QNS result as a function of log-transformed
# fluid volume, adjusting for specimen type. Log-volume is used because:
#   - Volume spans multiple orders of magnitude (0.1–22 mL)
#   - The QNS risk reduction per mL is unlikely to be linear on the
#     raw scale (going from 1 to 2 mL probably matters more than 21 to 22)
#   - log-linear modeling captures diminishing-returns behavior naturally
#
# The odds ratio for log(volume) represents the multiplicative change in
# QNS odds for each doubling of volume (we'll report it on the per-doubling
# scale to be clinically interpretable).

# Build the modeling dataset
df_logreg <- df |>
  filter(record_category %in% c("valid", "qns"),
         !is.na(vol_val), vol_val > 0,
         specimen_type_std %in% c("Blood", "Cyst Fluid")) |>
  mutate(
    qns_flag         = as.integer(record_category == "qns"),
    log2_vol         = log2(vol_val),
    specimen_type_std = factor(specimen_type_std, levels = c("Cyst Fluid", "Blood"))
  )

# Fit two models: volume-only (unadjusted) and volume + specimen type (adjusted)
fit_unadj <- glm(qns_flag ~ log2_vol,
                 data = df_logreg, family = binomial())
fit_adj   <- glm(qns_flag ~ log2_vol + specimen_type_std,
                 data = df_logreg, family = binomial())

# Extract odds ratios and 95% CIs
or_row <- function(fit, label) {
  s   <- summary(fit)$coefficients
  ci  <- suppressMessages(confint(fit))
  vol_row <- which(rownames(s) == "log2_vol")
  if (length(vol_row) == 0)
    return(tibble(Model = label, `OR per doubling` = "—",
                  `95% CI` = "—", `p-value` = "—"))
  tibble(
    Model              = label,
    `OR per doubling`  = sprintf("%.2f", exp(s[vol_row, "Estimate"])),
    `95% CI`           = sprintf("[%.2f–%.2f]",
                                  exp(ci[vol_row, 1]), exp(ci[vol_row, 2])),
    `p-value`          = fmt_p(s[vol_row, "Pr(>|z|)"])
  )
}

logreg_summary <- bind_rows(
  or_row(fit_unadj, "Volume only (unadjusted)"),
  or_row(fit_adj,   "Volume + specimen type (adjusted)")
)

# Specimen type effect from the adjusted model (Blood vs Cyst Fluid reference)
spec_coef <- summary(fit_adj)$coefficients
spec_row  <- grep("specimen_type_std", rownames(spec_coef), value = TRUE)
spec_effect <- if (length(spec_row) == 1) {
  ci_spec <- suppressMessages(confint(fit_adj))[spec_row, , drop = FALSE]
  glue("Blood vs. Cyst Fluid (adjusted): ",
       "OR = {sprintf('%.2f', exp(spec_coef[spec_row, 'Estimate']))} ",
       "[{sprintf('%.2f–%.2f', exp(ci_spec[1]), exp(ci_spec[2]))}], ",
       "p = {fmt_p(spec_coef[spec_row, 'Pr(>|z|)'])}")
} else {
  "Specimen-type effect: unavailable"
}

gt_logreg <- logreg_summary |>
  gt() |>
  tab_header(
    title    = "Logistic regression: probability of QNS by fluid volume",
    subtitle = "Outcome: QNS (vs. Valid). Predictor: log₂(volume) ± specimen type"
  ) |>
  tab_source_note(paste0(
    "OR per doubling = multiplicative change in QNS odds per doubling of volume ",
    "(e.g., OR = 0.50 means QNS odds halve with each doubling of volume; ",
    "OR < 1 indicates protective effect).",
    " | ", spec_effect,
    " | Restricted to Blood and Cyst Fluid specimens with volume > 0.",
    " | n = ", nrow(df_logreg), " (",
    sum(df_logreg$qns_flag == 1), " QNS, ",
    sum(df_logreg$qns_flag == 0), " Valid)."
  )) |>
  tab_options(table.width = pct(95), column_labels.font.weight = "bold")

gt_logreg
```

```{r preanalytical-qns-logreg-plot}
#| label: preanalytical-qns-logreg-plot
#| fig-width: 8
#| fig-height: 5

# Predicted-probability plot: P(QNS) as a function of volume, by specimen type.
# Useful as a visual companion to the OR table — shows the actual probability
# curve a clinician would want to see when deciding on a minimum submission volume.

# Build a smooth grid of volumes spanning the observed range
vol_grid <- seq(log2(min(df_logreg$vol_val)),
                log2(max(df_logreg$vol_val)),
                length.out = 200)

pred_df <- expand.grid(
  log2_vol         = vol_grid,
  specimen_type_std = factor(c("Cyst Fluid", "Blood"),
                              levels = c("Cyst Fluid", "Blood"))
)
pred_df$vol_val <- 2 ^ pred_df$log2_vol

# Predict with SE on the link scale, then back-transform
pred_link <- predict(fit_adj, newdata = pred_df, type = "link", se.fit = TRUE)
pred_df$prob    <- plogis(pred_link$fit)
pred_df$prob_lo <- plogis(pred_link$fit - 1.96 * pred_link$se.fit)
pred_df$prob_hi <- plogis(pred_link$fit + 1.96 * pred_link$se.fit)

p_qns_logreg_curve <- ggplot() +
  geom_ribbon(data = pred_df,
              aes(x = vol_val, ymin = prob_lo, ymax = prob_hi,
                  fill = specimen_type_std),
              alpha = 0.18) +
  geom_line(data = pred_df,
            aes(x = vol_val, y = prob, colour = specimen_type_std),
            linewidth = 1.1) +
  geom_jitter(data = df_logreg,
              aes(x = vol_val, y = qns_flag, colour = specimen_type_std),
              width = 0, height = 0.04, alpha = 0.5, size = 1.6) +
  scale_x_log10() +
  scale_y_continuous(labels = scales::percent,
                     breaks = seq(0, 1, 0.25), limits = c(-0.06, 1.06)) +
  scale_colour_manual(values = c("Cyst Fluid" = "#2C5282", "Blood" = "#C53030")) +
  scale_fill_manual(values   = c("Cyst Fluid" = "#2C5282", "Blood" = "#C53030")) +
  labs(
    x        = "Fluid volume (mL, log scale)",
    y        = "Predicted P(QNS)",
    colour   = "Specimen Type",
    fill     = "Specimen Type",
    title    = "Predicted probability of QNS as a function of fluid volume",
    subtitle = "Logistic regression: log₂(volume) + specimen type. Shaded bands = 95% CI."
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "top")

p_qns_logreg_curve
```

---

# Data Accounting

```{r data-accounting-summary}
#| label: data-accounting-summary

tibble(
  Item  = c(
    "Total rows in raw file",
    "Spacer rows excluded (no Patient ID)",
    "Records retained for analysis",
    "Unique patients"
  ),
  Count = c(n_raw, n_spacer, n_records, n_pts)
) |>
  gt() |>
  tab_header(title = "Data Accounting Summary") |>
  tab_options(table.width = pct(40), column_labels.font.weight = "bold")
```

> **Missing data policy:** All records with a valid Patient ID are retained (N = `r n_records`). For each analysis, only rows with valid (non-QNS, non-cancelled, non-blank) results are included. Blank result entries (N = `r n_blanks`) are excluded from all counts. Age and sex N reflect records with non-missing values only.

```{r consort-numbers}
#| label: consort-numbers

# Consolidated table of all N values needed to build the CONSORT-style flow
# diagram in PowerPoint/Keynote. Pulls every N from the canonical variables
# computed elsewhere in the analysis so the table is always synchronized with
# the current data. Transcribe these numbers directly into the manuscript's
# Figure 1 (study flow diagram).

# Test-level breakdowns by assay
n_pik3ca_valid <- sum(str_detect(df_valid$assay_std, "PIK3CA|PIK"))
n_pik3ca_pos   <- sum(str_detect(df_valid$assay_std, "PIK3CA|PIK") & df_valid$is_pos)
n_pik3ca_neg   <- sum(str_detect(df_valid$assay_std, "PIK3CA|PIK") & df_valid$is_neg)
n_tek_valid    <- sum(str_detect(df_valid$assay_std, "TEK"))
n_tek_pos      <- sum(str_detect(df_valid$assay_std, "TEK") & df_valid$is_pos)
n_tek_neg      <- sum(str_detect(df_valid$assay_std, "TEK") & df_valid$is_neg)
n_braf_valid   <- sum(str_detect(df_valid$assay_std, "BRAF"))
n_braf_pos     <- sum(str_detect(df_valid$assay_std, "BRAF") & df_valid$is_pos)
n_braf_neg     <- sum(str_detect(df_valid$assay_std, "BRAF") & df_valid$is_neg)

# Specimen type breakdowns (valid tests)
n_spec_blood   <- sum(df_valid$specimen_type_std == "Blood",      na.rm = TRUE)
n_spec_cyst    <- sum(df_valid$specimen_type_std == "Cyst Fluid", na.rm = TRUE)
n_spec_other   <- sum(df_valid$specimen_type_std == "Other",      na.rm = TRUE)

# Internal vs external (patient-level, all patients)
n_pts_internal <- sum(toupper(trimws(df_patients$internal_val)) == "Y", na.rm = TRUE)
n_pts_external <- sum(toupper(trimws(df_patients$internal_val)) == "N", na.rm = TRUE)

consort_df <- tribble(
  ~Stage, ~N,
  "STAGE 1 — INTAKE",                                                   NA_integer_,
  "  Total SpotSeq test records logged (Jan 2023 – Jan 2026)",          n_raw,
  "  Rows excluded: no Patient ID (spacer rows)",                       n_spacer,
  "  Records with valid Patient ID",                                    n_records,
  "  Unique patients",                                                  n_pts,
  "",                                                                   NA_integer_,
  "STAGE 2 — RESULT ELIGIBILITY",                                       NA_integer_,
  "  Records excluded: blank result entry",                             n_blanks,
  "  Records included in analysis",                                     n_total,
  "",                                                                   NA_integer_,
  "STAGE 3 — TEST VALIDITY",                                            NA_integer_,
  "  Cancelled tests (excluded from result analysis)",                  n_can,
  "  QNS — quantity not sufficient (excluded from result analysis)",    n_qns_n,
  "  Valid (analyzable: Positive or Negative)",                         n_valid,
  "    Positive results",                                               n_pos_all,
  "    Negative results",                                               n_neg_all,
  "",                                                                   NA_integer_,
  "STAGE 4 — VALID TESTS BY ASSAY",                                     NA_integer_,
  "  PIK3CA Multiplex",                                                 n_pik3ca_valid,
  "    Positive",                                                       n_pik3ca_pos,
  "    Negative",                                                       n_pik3ca_neg,
  "  TEK",                                                              n_tek_valid,
  "    Positive",                                                       n_tek_pos,
  "    Negative",                                                       n_tek_neg,
  "  BRAF",                                                             n_braf_valid,
  "    Positive",                                                       n_braf_pos,
  "    Negative",                                                       n_braf_neg,
  "",                                                                   NA_integer_,
  "STAGE 5 — VALID TESTS BY SPECIMEN TYPE",                             NA_integer_,
  "  Blood",                                                            n_spec_blood,
  "  Cyst Fluid",                                                       n_spec_cyst,
  "  Other (FFPE Scrolls / non-fluid)",                                 n_spec_other,
  "",                                                                   NA_integer_,
  "STAGE 6 — PATIENT ORIGIN (patient-level, all patients)",             NA_integer_,
  "  Internally processed at SCH",                                      n_pts_internal,
  "  Externally referred",                                              n_pts_external
)

# Render as gt for HTML display — display "" instead of NA for section headers
gt_consort <- consort_df |>
  mutate(N = if_else(is.na(N), "", as.character(N))) |>
  gt() |>
  tab_header(
    title    = "CONSORT-style Flow Diagram: Consolidated N Values",
    subtitle = "Transcribe these numbers into the manuscript Figure 1 flow diagram"
  ) |>
  tab_style(
    style     = list(cell_fill(color = "#D9E1F2"), cell_text(weight = "bold")),
    locations = cells_body(rows = grepl("^STAGE", consort_df$Stage))
  ) |>
  tab_source_note(paste0(
    "All N values are derived from canonical analysis variables and update ",
    "automatically with each render. Blank rows are visual separators between stages."
  )) |>
  tab_options(table.width = pct(80), column_labels.font.weight = "bold")

gt_consort
```

---

# Data Diagnostics — Unrecognised Categories

```{r diagnostics}
#| label: diagnostics
#| results: asis

# Surface anything in the new data that the analysis does NOT explicitly handle,
# so categories aren't silently dropped or miscounted.

diag_assays <- df_valid |>
  filter(!str_detect(assay_std, "PIK3CA|PIK|TEK|BRAF|GNAQ")) |>
  count(assay_std, sort = TRUE)

diag_specimens <- df_valid |>
  filter(specimen_type_std == "Other") |>
  count(specimen_type, sort = TRUE)

canon_re <- "E542K|GLU542LYS|1624G>A|E545K|GLU545LYS|1633G>A|H1047R|HIS1047ARG|3140A>G"
diag_variants <- df_valid |>
  filter(is_pos, str_detect(assay_std, "PIK3CA|PIK")) |>
  filter(!grepl(canon_re, variant_val, ignore.case = TRUE) &
         !grepl(canon_re, gene_val,    ignore.case = TRUE)) |>
  count(gene_val, variant_val, sort = TRUE)

clin_cols <- c("any_documented_management_change_in_management_post_testing_y_n",
               "targeted_therapy_initiated_post_testing_e_g_alpelisib_sirolimus_y_n",
               "existing_therapy_modified_or_dose_adjusted_post_testing_y_n_na",
               "surgical_or_procedural_plan_changed_post_testing_y_n_na",
               "patient_underwent_additional_testing_based_on_spot_seq_results_y_n_van",
               "management_documentation_post_testing_unavailable_not_reported_y_n_na")
missing_clin <- setdiff(clin_cols, names(df_raw))

if (nrow(diag_assays) + nrow(diag_specimens) + nrow(diag_variants) +
    length(missing_clin) == 0) {
  cat("No unrecognised categories detected. All valid tests are handled by the analysis.\n")
}
```

```{r diagnostics-tables}
#| label: diagnostics-tables

if (nrow(diag_assays) > 0) {
  diag_assays |> gt() |>
    tab_header(title = "Unhandled Assays",
               subtitle = "These tests are NOT counted in PIK3CA/TEK/BRAF rows of Tables 2/3.") |>
    tab_options(table.width = pct(50))
}
```

```{r diagnostics-specimens}
#| label: diagnostics-specimens

if (nrow(diag_specimens) > 0) {
  diag_specimens |> gt() |>
    tab_header(title = "Specimen Types Grouped Under 'Other'") |>
    tab_options(table.width = pct(50))
}
```

```{r diagnostics-variants}
#| label: diagnostics-variants

if (nrow(diag_variants) > 0) {
  diag_variants |> gt() |>
    tab_header(title = "PIK3CA-Positive Tests with Non-Canonical Variants",
               subtitle = "Counted in PIK3CA Positive total, but NOT in any sub-row.") |>
    tab_options(table.width = pct(50))
}
```

```{r diagnostics-clincols}
#| label: diagnostics-clincols
#| results: asis

if (length(missing_clin) > 0) {
  cat("**WARNING — clinical-impact columns missing (Table 4 may be wrong):**\n\n")
  cat(paste("-", missing_clin, collapse = "\n"))
  cat("\n")
}
```

```{r diagnostics-blank-rows}
#| label: diagnostics-blank-rows

# Show every record that was excluded as "blank" — has Patient ID but no
# parseable result. Useful for finding rows where the result column needs
# correction in the source spreadsheet.
blank_rows <- df |>
  filter(record_category == "blank") |>
  select(any_of(c("patient_id", "sample_id_for_di_dataset", "assay_ordered",
                  "result_pos_neg", "final_result", "status",
                  "specimen_type", "collection_date")))

if (nrow(blank_rows) == 0) {
  cat("✅ No blank result entries detected.\n")
} else {
  cat("⚠️ Records flagged as blank (excluded from analysis):\n\n")
  knitr::kable(blank_rows)
}
```

```{r diagnostics-vaf}
#| label: diagnostics-vaf

# Distribution check on parsed Internal VAF values.
# Surfaces values that are biologically implausible (>100% or =0 for Positives)
# so the user can investigate likely data entry / formatting errors.

if (!is.null(col_internal_vaf)) {
  vaf_summary <- df |>
    mutate(raw = .data[[col_internal_vaf]]) |>
    select(any_of(c("patient_id", "sample_id_for_di_dataset", "assay_ordered",
                    "result_pos_neg")),
           raw, internal_vaf_val, record_category)

  # Pos-only summary stats
  vaf_pos <- vaf_summary |>
    filter(record_category == "valid", grepl("^POS",
            toupper(trimws(.data[["result_pos_neg"]])))) |>
    pull(internal_vaf_val)
  vaf_pos <- vaf_pos[!is.na(vaf_pos)]

  cat("**Parsed VAF (positive valid results):**\n\n")
  if (length(vaf_pos) == 0) {
    cat("- No positive results with parseable VAF found.\n")
  } else {
    cat(sprintf("- N = %d\n", length(vaf_pos)))
    cat(sprintf("- Min: %.3f%%\n",  min(vaf_pos)))
    cat(sprintf("- Median: %.3f%%\n", median(vaf_pos)))
    cat(sprintf("- Mean: %.3f%%\n", mean(vaf_pos)))
    cat(sprintf("- Max: %.3f%%\n",  max(vaf_pos)))
  }
  cat("\n")

  # Flag implausible values (>100% or 0 for positives)
  flag <- vaf_summary |>
    filter(!is.na(internal_vaf_val) &
           (internal_vaf_val > 100 |
            (internal_vaf_val == 0 &
             grepl("^POS", toupper(trimws(.data[["result_pos_neg"]]))))))
  if (nrow(flag) > 0) {
    cat("**⚠️ VAF values flagged as biologically implausible** ",
        "(>100% suggests Excel formatting/data entry error; ",
        "0% on a Positive result suggests missing data):\n\n")
    knitr::kable(flag |> select(any_of(c("patient_id",
      "sample_id_for_di_dataset", "assay_ordered", "result_pos_neg",
      "raw", "internal_vaf_val"))))
  } else {
    cat("✅ No VAF values flagged as implausible.\n")
  }
} else {
  cat("⚠️ Internal VAF column not found in source dataset.\n")
}
```

```{r diagnostics-unmatched-spec-source}
#| label: diagnostics-unmatched-spec-source

# Surface specimen_source values that did not match Blood / Lymphatic Fluid /
# Other categories — these get silently excluded from the Specimen Source
# section of Table 2 and may indicate new descriptors that need adding to
# the predefined mapping.
unmatched_src <- df_valid |>
  filter(is.na(specimen_source_std)) |>
  count(specimen_source = .data[["specimen_source"]], sort = TRUE)

if (nrow(unmatched_src) == 0) {
  cat("✅ All valid-test specimen sources matched a defined category.\n")
} else {
  cat("⚠️ Specimen source values NOT matched to any category ",
      "(silently excluded from Specimen Source rows in Table 2):\n\n")
  knitr::kable(unmatched_src)
}
```

<!-- HTML stays local. Use `quarto render ... -P export:true` to generate Excel + Word + Drive upload. -->

---

```{r export-setup}
#| label: export-setup
#| eval: !expr params$export

# Export-only libraries — loaded only when params$export = TRUE
library(openxlsx)
library(flextable)
library(officer)
library(googledrive)

drive_folder_id <- Sys.getenv("DRIVE_FOLDER_ID")
stopifnot("DRIVE_FOLDER_ID not set in .Renviron" = nchar(drive_folder_id) > 0)

date_str   <- as.character(Sys.Date())
# Outputs folder is project-relative (Quarto sets wd to .qmd location).
# Tracked in Git so each export is version-controlled alongside the code.
output_dir <- "outputs"
dir.create(output_dir, showWarnings = FALSE)
```

```{r export-flextables}
#| label: export-flextables
#| eval: !expr params$export

# Generic flextable styler — section + p-value rows + footnote
ft_style <- function(ft, section_rows = integer(0), pval_rows = integer(0),
                     footnote = NULL, font_size = 10) {
  ft <- ft |>
    bold(part = "header") |>
    fontsize(size = font_size, part = "all") |>
    font(fontname = "Calibri", part = "all") |>
    border_outer(part = "all", border = fp_border(color = "black",   width = 1)) |>
    border_inner_h(part = "body", border = fp_border(color = "#CCCCCC", width = 0.5)) |>
    autofit()
  if (length(section_rows) > 0) {
    ft <- ft |>
      bg(i = section_rows, bg = "#D9E1F2", part = "body") |>
      bold(i = section_rows, part = "body")
  }
  if (length(pval_rows) > 0) {
    ft <- ft |>
      italic(i = pval_rows, part = "body") |>
      color(i = pval_rows, color = "#595959", part = "body")
  }
  if (!is.null(footnote)) {
    ft <- ft |>
      add_footer_lines(footnote) |>
      fontsize(size = 8, part = "footer") |>
      font(fontname = "Calibri", part = "footer") |>
      color(color = "#595959", part = "footer")
  }
  ft
}

ft_t1 <- flextable(t1) |>
  add_header_lines(glue(
    "Table 1. Patient and Cohort Characteristics (N = {n_records} records; {n_pts} unique patients)")) |>
  ft_style(
    section_rows = section_rows,
    footnote = glue("SD = standard deviation. IQR = interquartile range. Age from column AP (days) / 365.25. Patient-level rows computed across all unique patients (N = {n_pts}); subgroup denominators exclude missing values. Internally processed tests = test-level count.")
  )

ft_t2 <- flextable(t2_df) |>
  add_header_lines(glue("Table 2. Test Volume, Validity, and Results (N = {n_total} tests)")) |>
  ft_style(
    section_rows = section2_rows,
    footnote = "QNS = quantity not sufficient. Blank result entries excluded from all counts. NEG* treated as Negative. Fluid volume reported at patient level (unique patient ID); DNA and droplet metrics at test level (all valid tests)."
  )

ft_t3 <- flextable(table3_df) |>
  add_header_lines("Table 3. Assay Performance Metrics by Result (valid tests only)") |>
  ft_style(
    section_rows = section3_rows,
    pval_rows    = pval3_rows,
    font_size    = 8,
    footnote = paste(
      "IQR = interquartile range. VAF = variant allele frequency.",
      "Continuous metrics reported as Median [IQR]; VAF reported as mean (n contributing).",
      "Specimen type breakdown reflects tests within each result group.",
      "p-values: Mann-Whitney U test for continuous variables;",
      "Fisher's exact test for specimen type (Blood vs Cyst Fluid).",
      "VAF not statistically compared (negatives have no detected variant by definition).",
      "PIK3CA variant sub-row percentages are of PIK3CA Positive tests."
    )
  )

ft_t4 <- flextable(t4) |>
  add_header_lines(glue(
    "Table 4. Clinical Impact of Test Results (Total patients assessed, N = {n_t4_total})")) |>
  ft_style(
    section_rows = section4_rows,
    footnote = "Clinical actions were assessed in internally managed patients with valid (non-QNS, non-cancelled) results, as complete clinical documentation was available only for this subgroup. Y and VAN counted as positive for additional testing. Blank entries excluded from denominators. Targeted therapies listed are examples; actual agents confirmed by chart review."
  )
```

```{r export-word-docx}
#| label: export-word-docx
#| eval: !expr params$export

save_docx <- function(ft, filename, landscape = FALSE) {
  doc <- read_docx()
  if (landscape) {
    doc <- doc |>
      body_add_par("") |> body_end_section_continuous() |>
      body_add_flextable(ft) |> body_end_section_landscape()
  } else {
    doc <- doc |> body_add_flextable(ft)
  }
  print(doc, target = file.path(output_dir, filename))
  cat("✅ Saved:", filename, "\n")
}

save_docx(ft_t1, paste0("Table1_", date_str, ".docx"))
save_docx(ft_t2, paste0("Table2_", date_str, ".docx"))
save_docx(ft_t3, paste0("Table3_", date_str, ".docx"), landscape = TRUE)
save_docx(ft_t4, paste0("Table4_", date_str, ".docx"))

cat("\n✅ Word tables saved to:", output_dir, "\n")
```

```{r export-supp-tables}
#| label: export-supp-tables
#| eval: !expr params$export

# Supplementary tables — flextable docx exports for the pre-analytical
# correlation analyses and the CONSORT N table. Each is saved as a separate
# .docx file in outputs/ so it can be opened and pasted into the manuscript.

# --- Spearman correlations table ---
ft_spearman <- flextable(corr_results) |>
  add_header_lines(
    "Supplementary Table. Spearman correlations: fluid volume vs. wild-type droplet counts") |>
  ft_style(
    footnote = paste(
      "PIK3CA Multiplex valid tests, stratified by specimen type.",
      "Spearman rank correlation (non-parametric, distribution-free).",
      "Restricted to specimens with fluid volume > 0. Correlations not reported where n < 4."
    )
  )

# --- QNS rate by volume quartile table ---
ft_quartile <- flextable(quartile_summary) |>
  add_header_lines(glue(
    "Supplementary Table. QNS rate by fluid-volume quartile ",
    "(Cochran–Armitage trend test: p = {trend_p})")) |>
  ft_style(
    footnote = paste(
      "Quartiles defined on the combined valid + QNS pool (n ≈ equal per bin).",
      "Trend test assesses whether QNS proportion decreases monotonically with",
      "increasing volume. Useful for setting a clinical minimum-volume recommendation."
    )
  )

# --- Specimen-type-stratified QNS analysis table ---
ft_qns_spec <- flextable(qns_spec_summary) |>
  add_header_lines(
    "Supplementary Table. Fluid volume — QNS vs. valid, by specimen type") |>
  add_header_lines(paste(spec_pvalues, collapse = " | ")) |>
  ft_style(
    footnote = paste(
      "Two-sided Mann–Whitney U test within each specimen type.",
      "Restricted to specimens with fluid volume > 0.",
      "Stratified analysis avoids confounding by specimen-type-dependent",
      "differences in expected volume and collection workflow."
    )
  )

# --- Logistic regression results table ---
ft_logreg <- flextable(logreg_summary) |>
  add_header_lines(
    "Supplementary Table. Logistic regression: probability of QNS by fluid volume") |>
  ft_style(
    footnote = paste0(
      "OR per doubling = multiplicative change in QNS odds per doubling of volume ",
      "(OR < 1 indicates protective effect). ",
      spec_effect,
      ". Restricted to Blood and Cyst Fluid specimens with volume > 0. ",
      "n = ", nrow(df_logreg), " (",
      sum(df_logreg$qns_flag == 1), " QNS, ",
      sum(df_logreg$qns_flag == 0), " Valid)."
    )
  )

# --- CONSORT N table ---
# Convert NA → "" for cleaner display in Word
consort_for_export <- consort_df |>
  mutate(N = if_else(is.na(N), "", as.character(N)))
ft_consort <- flextable(consort_for_export) |>
  add_header_lines("CONSORT Flow Diagram: Consolidated N Values") |>
  bold(i = grep("^STAGE", consort_for_export$Stage), part = "body") |>
  bg(i = grep("^STAGE", consort_for_export$Stage),
     bg = "#D9E1F2", part = "body") |>
  fontsize(size = 10, part = "all") |>
  font(fontname = "Calibri", part = "all") |>
  border_outer(part = "all", border = fp_border(color = "black",   width = 1)) |>
  border_inner_h(part = "body", border = fp_border(color = "#CCCCCC", width = 0.5)) |>
  add_footer_lines(paste(
    "All N values are derived from canonical analysis variables.",
    "Transcribe directly into the manuscript Figure 1 flow diagram."
  )) |>
  fontsize(size = 8, part = "footer") |>
  color(color = "#595959", part = "footer") |>
  autofit()

# Save each as a separate docx
save_docx(ft_spearman,   paste0("Supp_Spearman_",       date_str, ".docx"))
save_docx(ft_quartile,   paste0("Supp_QNSQuartile_",    date_str, ".docx"))
save_docx(ft_qns_spec,   paste0("Supp_QNSbySpecType_",  date_str, ".docx"))
save_docx(ft_logreg,     paste0("Supp_QNSLogReg_",      date_str, ".docx"))
save_docx(ft_consort,    paste0("Supp_CONSORTcounts_",  date_str, ".docx"))

cat("\n✅ Supplementary tables saved to:", output_dir, "\n")
```

```{r export-supp-plots}
#| label: export-supp-plots
#| eval: !expr params$export

# Supplementary plots — saved as high-resolution PNG (300 dpi) for direct
# insertion into the Word manuscript. Also saves PDF versions for journal
# submission (most journals prefer vector graphics for figures).

save_plot <- function(plot_obj, basename, w, h) {
  png_path <- file.path(output_dir, paste0(basename, "_", date_str, ".png"))
  pdf_path <- file.path(output_dir, paste0(basename, "_", date_str, ".pdf"))
  ggsave(png_path, plot = plot_obj, width = w, height = h, dpi = 300, bg = "white")
  ggsave(pdf_path, plot = plot_obj, width = w, height = h,           bg = "white")
  cat("✅ Saved:", basename, "(PNG + PDF)\n")
}

save_plot(p_spearman_scatter, "Fig_SpearmanScatter",  w = 9, h = 7)
save_plot(p_qns_spectype,     "Fig_QNSbySpecType",    w = 8, h = 4.5)
save_plot(p_qns_logreg_curve, "Fig_QNSLogRegCurve",   w = 8, h = 5)

cat("\n✅ Supplementary plots saved to:", output_dir, "\n")
```

```{r export-excel}
#| label: export-excel
#| eval: !expr params$export

wb       <- createWorkbook()
title_st <- createStyle(textDecoration = "bold", fontSize = 13)
hdr_st   <- createStyle(textDecoration = "bold", fontSize = 12)
sec_st   <- createStyle(textDecoration = "bold", fgFill = "#D9E1F2")
ital_st  <- createStyle(textDecoration = "italic", fontSize = 10, fontColour = "#595959")
norm_st  <- createStyle(fontSize = 11)
fn_st    <- createStyle(fontSize = 9, fontColour = "#595959", wrapText = TRUE)

write_table_sheet <- function(sheet_name, title, df_data,
                              section_rows = integer(0),
                              pval_rows    = integer(0),
                              footnote     = NULL,
                              col_widths   = NULL) {
  addWorksheet(wb, sheet_name)
  ncols <- ncol(df_data)
  if (!is.null(col_widths))
    setColWidths(wb, sheet_name, cols = 1:ncols, widths = col_widths)

  # Title row
  writeData(wb, sheet_name, title, startRow = 1, startCol = 1, colNames = FALSE)
  addStyle(wb, sheet_name, title_st, rows = 1, cols = 1)
  mergeCells(wb, sheet_name, 1:ncols, 1)

  # Header + body data (writeData puts headers at startRow, data at startRow+1)
  writeData(wb, sheet_name, df_data, startRow = 3, headerStyle = hdr_st)

  # Per-row styling — body rows start at Excel row 4
  body_start <- 4
  for (i in seq_len(nrow(df_data))) {
    r  <- body_start + i - 1
    st <- if (i %in% pval_rows)        ital_st
          else if (i %in% section_rows) sec_st
          else                          norm_st
    addStyle(wb, sheet_name, st, rows = r, cols = 1:ncols, gridExpand = TRUE)
  }

  # Footnote
  if (!is.null(footnote)) {
    fn_row <- body_start + nrow(df_data) + 1
    writeData(wb, sheet_name, footnote, startRow = fn_row, startCol = 1, colNames = FALSE)
    addStyle(wb, sheet_name, fn_st, rows = fn_row, cols = 1)
    mergeCells(wb, sheet_name, 1:ncols, fn_row)
  }
}

write_table_sheet(
  "Table 1",
  glue("Table 1. Patient and Cohort Characteristics (N={n_records}; {n_pts} unique patients)"),
  t1, section_rows = section_rows, col_widths = c(50, 25),
  footnote = glue("SD = standard deviation. IQR = interquartile range. Age from column AP (days) / 365.25. Patient-level rows computed across all unique patients (N = {n_pts}); subgroup denominators exclude missing values. Internally processed tests = test-level count.")
)

write_table_sheet(
  "Table 2",
  glue("Table 2. Test Volume, Validity, and Results (N={n_total} tests)"),
  t2_df, section_rows = section2_rows, col_widths = c(44, 14, 18, 18),
  footnote = "QNS = quantity not sufficient. Blank result entries excluded from all counts. NEG* treated as Negative."
)

write_table_sheet(
  "Table 3",
  "Table 3. Assay Performance Metrics by Result (valid tests only)",
  table3_df, section_rows = section3_rows, pval_rows = pval3_rows,
  col_widths = c(20, 18, 10, 18, 20, 18, 22, 16, 20),
  footnote = paste(
    "IQR = interquartile range. VAF = variant allele frequency.",
    "Continuous metrics reported as Median [IQR]; VAF reported as mean (n contributing).",
    "Specimen type breakdown reflects tests within each result group.",
    "p-values: Mann-Whitney U test for continuous variables; Fisher's exact test for specimen type (Blood vs Cyst Fluid).",
    "VAF not statistically compared (negatives have no detected variant by definition).",
    "PIK3CA variant sub-row percentages are of PIK3CA Positive tests."
  )
)

write_table_sheet(
  "Table 4",
  glue("Table 4. Clinical Impact of Test Results (Total patients assessed, N = {n_t4_total})"),
  t4, section_rows = section4_rows, col_widths = c(55, 18),
  footnote = "Y and VAN counted as positive for additional testing. Blank entries excluded from denominators."
)

xlsx_path <- file.path(output_dir, paste0("SpotSeq_Tables_", date_str, ".xlsx"))
saveWorkbook(wb, xlsx_path, overwrite = TRUE)
cat("✅ Excel saved:", xlsx_path, "\n")
```

```{r export-drive-upload}
#| label: export-drive-upload
#| eval: !expr params$export

drive_auth(
  scopes = "https://www.googleapis.com/auth/drive",
  cache  = "~/.gargle-cache",
  email  = "solomonj88@gmail.com"
)
drive_upload(
  media     = xlsx_path,
  path      = as_id(drive_folder_id),
  name      = paste0("SpotSeq_Tables_", date_str, ".xlsx"),
  overwrite = TRUE
)
cat("✅ Uploaded to Google Drive folder:", drive_folder_id, "\n")
cat("\n🔖 GitHub checkpoint:\n")
cat('  git add . && git commit -m "feat: consolidated QMD-only analysis with parameterised exports"\n')
```
