---
title: "SpotSeq Dataset Analysis"
author: "Your Name"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    code-fold: true
    code-tools: true
    embed-resources: true
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
---

# SpotSeq Dataset Analysis Setup

This document is the master project skeleton for the SpotSeq dataset analysis.
It handles environment setup, SharePoint/OneDrive data connection, Google Drive
output configuration, and GitHub version control checkpoints — all in one place.

---

## Step 1: Package Setup

Run this chunk once to install and snapshot all required packages.

```{r}
#| label: install-packages
#| eval: false  # Set to true on first run only

# Install required packages
install.packages(c(
  "renv",           # Package version control
  "googledrive",    # Push outputs to Google Drive
  "Microsoft365R",  # Connect to SharePoint / OneDrive
  "tidyverse",      # Core data wrangling
  "here",           # Robust file path handling
  "janitor",        # Data cleaning utilities
  "skimr",          # Quick data summaries
  "quarto"          # Quarto rendering from R
))

# Snapshot the environment to renv.lock
renv::snapshot()
```

> **Note:** After your first run, set `eval: false` above and never change it again.
> Your `renv.lock` file in GitHub will handle reproducibility from here on.

---

## Step 2: Load Libraries

```{r}
#| label: load-libraries

library(tidyverse)
library(googledrive)
library(Microsoft365R)
library(here)
library(janitor)
library(skimr)
```

---

## Step 3: Configure Your Environment

Run this chunk **once** to write your project paths into `.Renviron`.
After running it, restart R — you will never need to hardcode paths again.

```{r}
#| label: configure-renviron
#| eval: false  # Set to true on first run only, then back to false

# Opens your project .Renviron file for editing
usethis::edit_r_environ(scope = "project")

# Paste the following lines into the file that opens, then save and restart R:
#
# SHAREPOINT_SITE = "https://yourorg.sharepoint.com/sites/your-site"
# SHAREPOINT_FOLDER = "Shared Documents/SpotSeq/data"
# DRIVE_FOLDER_ID  = "paste-your-google-drive-folder-id-here"
#
# The Drive folder ID is printed at the end of Step 5 below.
```

---

## Step 4: Authenticate Connections

```{r}
#| label: authenticate
#| eval: false  # Run interactively once per session as needed

# Authenticate Google Drive (opens browser on first run)
googledrive::drive_auth()

# Authenticate SharePoint / OneDrive (opens browser on first run)
sharepoint <- Microsoft365R::get_sharepoint_site(
  Sys.getenv("SHAREPOINT_SITE")
)
```

---

## Step 5: Create Google Drive Output Folders

Run this chunk **once** at project start to build your Drive folder structure.

```{r}
#| label: create-drive-folders
#| eval: false  # Run once, then set back to false

library(googledrive)

# Create top-level project folder
root <- drive_mkdir("SpotSeq-Dataset-Analysis")

# Create subfolders for outputs
subfolders <- c("reports", "figures", "outputs")
sapply(subfolders, function(f) drive_mkdir(f, path = root))

# Print the root folder ID — copy this into your .Renviron as DRIVE_FOLDER_ID
cat("\n Drive root folder ID (copy to .Renviron):\n", root$id, "\n")
```

---

## Step 6: Connect to SharePoint Data

```{r}
#| label: connect-sharepoint
#| eval: false  # Enable when ready to load data

# Connect to SharePoint
sharepoint <- Microsoft365R::get_sharepoint_site(
  Sys.getenv("SHAREPOINT_SITE")
)

# Browse available files (run interactively to explore)
sp_folder <- sharepoint$get_drive()$get_item(
  Sys.getenv("SHAREPOINT_FOLDER")
)

# List files in the folder
sp_folder$list_files()
```

```{r}
#| label: load-data
#| eval: false  # Enable when ready to load data

# Example: Download a specific file from SharePoint into R
# Replace "your_file.csv" with your actual filename

sp_folder$get_item("your_file.csv")$download(
  dest = here("data", "raw", "your_file.csv"),
  overwrite = TRUE
)

# Read into R
raw_data <- read_csv(
  here("data", "raw", "your_file.csv"),
  show_col_types = FALSE
) |>
  clean_names()

# Quick summary
glimpse(raw_data)
```

---

## Step 7: Push Outputs to Google Drive

Use this chunk after rendering to push your report to Drive.

```{r}
#| label: push-to-drive
#| eval: false  # Run after rendering

library(googledrive)

# Get Drive folder reference from stored ID
drive_folder <- as_id(Sys.getenv("DRIVE_FOLDER_ID"))

# Push the rendered HTML report
drive_upload(
  media    = here("outputs", "spotseq_analysis.html"),
  path     = drive_folder,
  name     = paste0("SpotSeq_Report_", Sys.Date(), ".html"),
  overwrite = TRUE
)

cat("Report pushed to Google Drive successfully.\n")
```

---

## Step 8: Git Version Control Checkpoints

Use these commands in your R terminal to commit progress to GitHub.

```{r}
#| label: git-checkpoint
#| eval: false  # Run manually at each milestone

# Stage all changes
system("git add .")

# Commit with a descriptive message — edit the message below
system('git commit -m "milestone: describe what you just completed"')

# Push to GitHub
system("git push")
```

**Suggested commit messages by milestone:**

| Milestone | Message |
|---|---|
| After setup | `"chore: initial project skeleton"` |
| After data load | `"feat: SharePoint data connection and raw load"` |
| After cleaning | `"feat: data cleaning pipeline complete"` |
| Before sharing | `"release: v1.0 initial analysis"` |

**To tag a release milestone:**

```{r}
#| label: git-tag
#| eval: false

system('git tag -a v1.0 -m "Initial SpotSeq analysis release"')
system("git push origin v1.0")
```

---

## Project Status Tracker

Update this table as you complete each step:

| Step | Description | Status |
|---|---|---|
| 1 | Packages installed & snapshotted | ⬜ Pending |
| 2 | Libraries loading cleanly | ⬜ Pending |
| 3 | `.Renviron` configured | ⬜ Pending |
| 4 | SharePoint + Drive authenticated | ⬜ Pending |
| 5 | Drive folders created | ⬜ Pending |
| 6 | Data loading from SharePoint | ⬜ Pending |
| 7 | Outputs pushing to Drive | ⬜ Pending |
| 8 | First GitHub commit pushed | ⬜ Pending |

---

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
